---
title: "Drosophila Burst Speed"
author: "Jeff Walker"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, echo=FALSE}
# wrangling packages
library(data.table) # magical data frames
library(here) # here makes a project transportable
library(janitor) # clean_names
library(readxl) # read_excel, duh!
library(writexl) # write_xlsx, duh!
library(stringr) # text wrangling

# analysis packages
library(lme4)
library(lmerTest)
library(emmeans) # the workhorse for inference


# graphing packages
library(ggsci) # color palettes
library(ggpubr) # publication quality plots
library(ggforce) # better jitter
library(cowplot) # combine plots
library(knitr) # kable tables
library(kableExtra) # kable_styling tables

# ggplot_the_model.R packages not loaded above
library(insight)
library(lazyWeave)
library(afex) # aov4

# use here from the here package
here <- here::here
# use clean_names from the janitor package
clean_names <- janitor::clean_names
# use transpose from data.table
transpose <- data.table::transpose

# load functions used by this text written by me
# ggplot_the_model.R needs to be in the folder "R"
# if you didn't download this and add to your R folder in your
# project, then this line will cause an error

source(here("code", "ggplot_the_model.R"))
source(here("code", "ggptm.R"))

data_folder <- "data"

# constants
dial_to_speed <- 6.42237212 # from my smoke test with screens and forcing through origin

```


# Burst analysis
## Experiment 1 (Simons-Walker)
### Import
The major analyis is an estimate of the effects.

```{r import}
subfolder <- "burst speed/simons-walker "
filename <- "Wind Tunnel minus days>6.xlsx"
in_path <- here(data_folder, subfolder, filename)

exp1 <- read_xlsx(in_path,
                    range = "A1:H108") |>
  clean_names() |>
  data.table()

exp1[, id := .I]
exp1[, treatment := factor(treatment, levels = c("CN", "AA"))]
exp1[, sex := factor(sex, levels = c("M", "F"))]
exp1[, line := paste0(treatment, line) |>
       factor(levels = c("CN1", "CN2", "AA1", "AA2"))]
exp1[, wind_speed := dial_50 * dial_to_speed]
exp1[, group := ifelse(treatment == "AA", "Selection", "No Selection") |> factor(levels=c("No Selection", "Selection"))]
```

```{r exp1-checks}
exp1[, .(max = max(wind_speed)), by = .(line, sex)]
```

```{r exp1-init-plot}
ggplot(data = exp1,
       aes(x = line,
           y = wind_speed)) +
  geom_jitter(width = 0.1)
```


```{r exp1-sample-means}

exp1_summary <- exp1[, .(wind_speed = mean(wind_speed)), by = .(treatment, sex, line)]
exp1_summary
```

```{r exp1-room-temp}
lm_ancova <- lm(wind_speed ~ treatment * sex + room_temp, data = exp1)
coef(summary(lm_ancova))

ggplot(data = exp1,
       aes(x = room_temp,
           y = wind_speed,
           color = treatment)) +
  geom_point() +
#  geom_ancova(lm_ancova) +
  NULL
```

```{r exp1-balance}
exp1[, .(N = .N), by = .(line, sex)]
```
### Estimate Treatment Effects

```{r exp1-lmm-random-intercept}
exp1_lmm <- lmer(wind_speed ~ treatment * sex + (1 | line), data = exp1)
coef(summary(exp1_lmm))
sigma <- summary(exp1_lmm)$sigma

exp1_emm <- emmeans(exp1_lmm, specs = c("treatment", "sex"))

cn_m <- c(1,0,0,0)
aa_m <- c(0,1,0,0)
cn_f <- c(0,0,1,0)
aa_f <- c(0,0,0,1)
exp1_contrast <- list(
  list(male = aa_m - cn_m),
  list(female = aa_f - cn_f)
)

exp1_pairs <- contrast(exp1_emm,
                      method = exp1_contrast,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
exp1_pairs[, effect_std := estimate / sigma]

exp1_emm
exp1_pairs

```

### Plot

1. Separate sex
2. plot each line as a separate X value, show the data
3. get line means from lmm

```{r exp1-conditional-means}
# compute conditional means
newdat <- expand.grid(
  line = levels(exp1$line),
  sex = levels(exp1$sex)
) |>
  data.table()
newdat[, treatment := ifelse(substr(newdat$line, 1, 2) == "AA", "AA", "CN")]
newdat[, fit := predict(exp1_lmm, newdata = newdat)]

# check with my own math
coefs <- coef(exp1_lmm)$line
mm_males <- matrix(c(1,0,0,0,
                       1,0,0,0,
                       1,1,0,0,
                       1,1,0,0), nrow=4)
mm_females <- matrix(c(1,0,1,0,
                       1,0,1,0,
                       1,1,1,1,
                       1,1,1,1), nrow=4)
male_means <- as.matrix(coefs) %*% mm_males |> diag()
female_means <- as.matrix(coefs) %*% mm_females |> diag()

# compare with raw means
fixed_means <- exp1[, .(speed = mean(wind_speed)),
                      by = .(sex, line, treatment)]
fixed_means

```

```{r exp1-conditional-mean-plot}

means_plot <- function(plot_sex = "F", title = "Female"){
  offset <- 0.15
  x_breaks <- c(1-offset, 1+offset, 2-offset, 2+offset)
  
  plot_data <- exp1[, .SD, .SDcols = c("treatment", "line", "sex", "wind_speed")]
  treatment_means <- exp1_emm |> 
    summary() |>
    data.table()
  treatment_means[treatment == "CN", x := 1]
  treatment_means[treatment == "AA", x := 2]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "CN1", x := 1 - offset]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "AA1", x := 2 - offset]
  plot_data[line == "AA2", x := 2 + offset]
  newdat[line == "CN1", x := 1 - offset]
  newdat[line == "CN2", x := 1 + offset]
  newdat[line == "AA1", x := 2 - offset]
  newdat[line == "AA2", x := 2 + offset]
  
  gg <- ggplot(data = plot_data[sex == plot_sex],
               aes(x = x,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    geom_jitter(width = 0.05,
                height = 0,
                color = "gray") +
    
    # line means
    geom_point(data = newdat[sex == plot_sex],
               aes(x = x,
                   y = fit),
               size = 2) +
    
    # treatment means and CI
    geom_point(data = treatment_means[sex == plot_sex],
               aes(x = x,
                   y = emmean),
               size = 3) +
    geom_errorbar(data = treatment_means[sex == plot_sex],
                  aes(x = x,
                      y = emmean,
                      ymin = lower.CL,
                      ymax = upper.CL),
                  width = 0) +
    
    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = pal_okabe_ito) +
    scale_x_continuous(breaks = x_breaks,
                       labels = c("CN1", "CN2", "AA1", "AA2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(90, 210),
                    xlim = c(0.7, 2.3)) +
    ggtitle(title)
  return(gg)
}

diff_plot <- function(){
  gg <- ggplot(data = )
}

gg_f <- means_plot(plot_sex = "F", title = "Female")
gg_m <- means_plot(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```

### Other models
With room_temp covariate (supplement)

```{r exp1-lmm-random-intercept}
exp1_lmm_2 <- lmer(wind_speed ~ treatment * sex + room_temp + (1 | line), data = exp1)
coef(summary(exp1_lmm_2))
sigma <- summary(exp1_lmm_2)$sigma

exp1_lmm_2_emm <- emmeans(exp1_lmm_2, specs = c("treatment", "sex"))

exp1_lmm_2_pairs <- contrast(exp1_lmm_2_emm,
                      method = exp1_contrast,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
exp1_lmm_2_pairs[, effect_std := estimate / sigma]

exp1_lmm_2_emm
exp1_lmm_2_pairs

```
```{r exp1-classical-nested-anova}
str(exp1)
m1 <- aov(wind_speed ~ treatment * sex + Error(line / sex),
          data = exp1)
summary(m1)

m1_male <- aov(wind_speed ~ treatment + Error(line),
          data = exp1)
```


## Experiment 2 (Wheeler) data
### Clean and Import
```{r exp2-clean-his-files}
subfolder <- "burst speed/jess wheeler/sent from jess/raw"
filename <- "02.14.10.xlsx"
in_path <- here(data_folder, subfolder, filename)

  file_i <- read_xlsx(in_path,
                    range = "a1:c1000") |>
    clean_names() |>
    data.table() |>
    na.omit()
  # angle is flight angle in degrees from 1,0
  # length is flight distance over 1 frame
  colnames(file_i)[1] <- "id"
  file_i[, treatment := substr(id,1,2)]
  file_i[, line := substr(id,1,3)]
  file_i[, sex := substr(id, 4,4)]
  file_i[, dial := substr(id, 5,6) |> as.integer()]
  file_i[, date := substr(id, 8,15)]
 
  check_n <- file_i[, .(N = .N),
     by = .(date, sex, treatment, line, id)]
  check_n[N >=6, id]

  check_id <- check_n[, .(N = .N), 
     by = .(date, sex, treatment, line)]
setorder(check_id, sex, treatment, line, date)

```

```{r exp2-import-cleaned}

subfolder <- "burst speed/jess wheeler/sent from jess/raw"
subfolder_path <- here(data_folder, subfolder)

files <- list.files(
  path = subfolder_path,
  pattern = "\\.xlsx$",
  full.names = TRUE
)

exp2 <- data.table(NULL)

for(i in 1:length(files)){
  in_path <- files[i]
  
  file_i <- read_xlsx(in_path,
                    range = "a1:c1000") |>
    clean_names() |>
    data.table() |>
    na.omit()
  # angle is flight angle in degrees from 1,0
  # length is flight distance over 1 frame
  colnames(file_i)[1] <- "id"
  file_i[, treatment := substr(id,1,2)]
  file_i[, line := substr(id,1,3) |> 
           toupper() |>
           factor()]
  file_i[, sex := substr(id, 4,4) |> 
           toupper() |>
           factor()]
  file_i[, dial := substr(id, 5,6) |> as.integer()]
  file_i[, date := substr(id, 8,15)]
  
  # performance measures
  file_i[, theta := ifelse(angle < 0,
                         -(angle + 180),
                         180-angle)]
  file_i[, wind_jess := 7.3888 * dial]
  file_i[, wind_jaw := 6.42237212 * dial]
  file_i[, disp.x := length * cos(theta * pi/180)]
  file_i[, disp.y := length * sin(theta * pi/180)]
  file_i[, wind.x := wind_jess / 30]
  file_i[, vel := sqrt((wind.x + disp.x)^2 + disp.y^2) * 30]
  
  exp2 <- rbind(exp2, file_i)
  
}

  # clean treatments and make factor
  exp2[treatment == "aa", treatment := "AA"]
  exp2[treatment == "cn", treatment := "CN"]
  # exp2[treatment == "k4", treatment := "K45"]
  # exp2[treatment == "n6", treatment := "N60"]
  exp2[treatment == "k4", treatment := "Wild"]
  exp2[treatment == "n6", treatment := "Wild"]
  exp2[, treatment := factor(treatment, levels = 
                                 c("CN", "Wild", "AA"))]
  exp2[, line := factor(line, 
                        levels = c("CN1", "CN2",
                                   "K45", "N60",
                                   "AA1", "AA2"))]
  exp2[treatment == "AA", group := "Selection"]
  exp2[treatment == "CN", group := "No Selection"]
  exp2[treatment == "Wild", group := "Wild"]
  exp2[, group := factor(group,
                         levels = c("No Selection",
                                    "Selection",
                                    "Wild"))]

```

```{r exp2-summary}
# this is just to compare counts with Jess' thesis table
# first compute max within each id
exp2_max_id <- exp2[, .(wind_max = mean(wind_jess),
                        vel_max = mean(vel),
                        N = .N),
     by = .(date, treatment, line, sex, id)]

exp2_sum <- exp2_max_id[, .(wind_max = mean(wind_max),
                            vel_max = mean(vel_max),
                    N = .N), 
     by = .(date, sex, treatment, line)]
setorder(exp2_sum, sex, treatment, line, date)
```

### Primary data set

To compare to Exp1 (Simons-Walker) results, I am using wind tunnel speed and not the computed speed of the fly. But I will analyze that in a supplement. Jess has 3 measures per fly; since these are at same tunnel speed, I just need to obtain the summary data as the primary data set.

```{r jess-primary-data}
exp2_primary <- exp2[, .(wind_speed = mean(wind_jess)), 
                     by = .(id, treatment, line, sex, group)]
```

### Estimate Treatment Effects

```{r exp2-lmm-random-intercept}
exp2_lmm <- lmer(wind_speed ~ treatment * sex + (1 | line), data = exp2_primary)
coef(summary(exp2_lmm))
sigma <- summary(exp2_lmm)$sigma

exp2_emm <- emmeans(exp2_lmm, specs = c("treatment", "sex"))

cn_f <- c(1,0,0,0,0,0)
wild_f <- c(0,1,0,0,0,0)
aa_f <- c(0,0,1,0,0,0)
cn_m <- c(0,0,0,1,0,0)
wild_m <- c(0,0,0,0,1,0)
aa_m <- c(0,0,0,0,0,1)

# contrasts of interest - this is all contrasts, just not m v f
# aa - cn # primary contrast
# aa - k45 # secondary contrast testing comparison with newer line
# aa - n60 # secondary contrast testing comparison with newer line
# cn - k45 # secondary contrast testing loss in cn
# cn - n60 # secondary contrast testing loss in cn
# n60 - k45 # n60 is july 2009 and k45 is july 2008

exp2_contrasts <- list(
  list(aa_cn_male = aa_m - cn_m),
  list(aa_wt_male = aa_m - wild_m),
  list(cn_wt_male = cn_m - wild_m),
  list(aa_cn_female = aa_f - cn_f),
  list(aa_wt_female = aa_f - wild_f),
  list(cn_wt_female = cn_f - wild_f)
)

exp2_pairs <- contrast(exp2_emm,
                      method = exp2_contrasts,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
exp2_pairs[, effect_std := estimate / sigma]

exp2_emm
exp2_pairs

```

### Plot

```{r exp2-conditional-means}
# compute conditional means
newdat <- expand.grid(
  line = levels(exp2_primary$line),
  sex = levels(exp2_primary$sex)
) |>
  data.table()
# newdat[, treatment := line]
# newdat[substr(line,1,2) == "aa", treatment := "aa"]
# newdat[substr(line,1,2) == "cn", treatment := "cn"]
newdat[, treatment := substr(line,1,2)]
newdat[treatment == "K4", treatment := "Wild"]
newdat[treatment == "N6", treatment := "Wild"]
newdat[, fit := predict(exp2_lmm, newdata = newdat)]

```

```{r exp2-conditional-mean-plot}

means_plot <- function(plot_sex = "F", title = "Female"){
  offset <- 0.25
  x_breaks <- c(1-offset, 1+offset, 2-offset, 2+offset,
                3-offset, 3+offset)
  
  plot_data <- exp2_primary[, .SD, .SDcols = c("treatment", "line", "sex", "wind_speed")]
  treatment_means <- exp2_emm |> 
    summary() |>
    data.table()
  treatment_means[treatment == "CN", x := 1]
  treatment_means[treatment == "Wild", x := 2]
  treatment_means[treatment == "AA", x := 3]
  plot_data[line == "CN1", x := 1 - offset]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "K45", x := 2 - offset]
  plot_data[line == "N60", x := 2 + offset]
  plot_data[line == "AA1", x := 3 - offset]
  plot_data[line == "AA2", x := 3 + offset]
  newdat[line == "CN1", x := 1 - offset]
  newdat[line == "CN2", x := 1 + offset]
  newdat[line == "K45", x := 2 - offset]
  newdat[line == "N60", x := 2 + offset]
  newdat[line == "AA1", x := 3 - offset]
  newdat[line == "AA2", x := 3 + offset]

  gg <- ggplot(data = plot_data[sex == plot_sex],
               aes(x = x,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    geom_jitter(width = 0.05,
                height = 0.5,
                size = 0.2,
                color = "gray") +

    # line means
    geom_point(data = newdat[sex == plot_sex],
               aes(x = x,
                   y = fit),
               size = 2) +
    
    # treatment means and CI
    geom_point(data = treatment_means[sex == plot_sex],
               aes(x = x,
                   y = emmean),
               size = 3) +
    geom_errorbar(data = treatment_means[sex == plot_sex],
                  aes(x = x,
                      y = emmean,
                      ymin = lower.CL,
                      ymax = upper.CL),
                  width = 0) +
    
    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = c(pal_okabe_ito[1],
                                  pal_okabe_ito[3],
                                  pal_okabe_ito[2])) +
    scale_x_continuous(breaks = x_breaks,
                       labels = c("CN\n1", "CN\n2",
                                  "Wild\nK45", "Wild\nN60",
                                   "AA\n1", "AA\n2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(0, 250),
                    xlim = c(0.7, 3.3)) +
    ggtitle(title)
  return(gg)
}

gg_f <- means_plot(plot_sex = "F", title = "Female")
gg_m <- means_plot(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```

```{r exp2-conditional-mean-plot-violin}

means_plot <- function(plot_sex = "F", title = "Female"){

  gg <- ggplot(data = exp2_primary[sex == plot_sex],
               aes(x = line,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    # geom_jitter(width = 0.05,
    #             height = 0,
    #             color = "gray") +

    # geom_violin() +
    
    geom_boxplot() +

    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = 
                         c(pal_okabe_ito[1], 
                           pal_okabe_ito[3], 
                           pal_okabe_ito[2])) +
    scale_x_discrete(labels = c("CN\n1", "CN\n2",
                                "Wild\nK45", "Wild\nN60",
                                "AA\n1", "AA\n2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(25, 225)) +
    ggtitle(title)
  return(gg)
}

gg_f <- means_plot(plot_sex = "F", title = "Female")
gg_m <- means_plot(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```


## Combined

```{r burst-combined-plot}
burst1 <- data.table(
  experiment = "Exp1",
  exp1[, .SD, .SDcols = c("treatment", "line", "sex", "group",  "wind_speed")])
burst2 <- data.table(
  experiment = "Exp2",
  exp2[, .SD, .SDcols = c("treatment", "line", "sex", "group", "wind_jess")])
setnames(burst2, old = "wind_jess", new = "wind_speed")
burst <- rbind(burst1, burst2)
burst[, exp_line := paste(experiment, line, sep="\n")]
burst[, exp_line := factor(exp_line, levels = c(
  "Exp1\nCN1", "Exp1\nCN2",
  "Exp1\nAA1", "Exp1\nAA2",
  "Exp2\nCN1", "Exp2\nCN2",
  "Exp2\nAA1", "Exp2\nAA2",
  "Exp2\nK45", "Exp2\nN60"
))]

combined_plot <- function(sex_in = "F", title = "Female"){
  gg <- ggplot(data = burst[sex == sex_in],
               aes(x = exp_line,
                   y = wind_speed,
                   color = group)) +
  geom_jitter(width = 0.2,
              height = 1,
              size = .2) +
  # clean plot
  ylab("Max Opposing Wind Speed (cm/s)") +
  scale_color_manual(values = pal_okabe_ito) +
  theme_pubr() +
  theme(legend.position="none",
        axis.title.x = element_blank()) +
  ggtitle(title)
  return(gg)
}

gg_f <- combined_plot(sex = "F", title = "Female")
gg_f
gg_m <- combined_plot(sex = "M", title = "Male")
gg_m
```

# Kinematic analysis


```{r free-flight-import}
subfolder <- "kinematics/free flight/processed"
filename <- "frequency data.xlsx"
in_path <- here(data_folder, subfolder, filename)

free_flight <- read_xlsx(in_path,
                   sheet = "frequency data") |>
  clean_names() |>
  data.table() |>
  na.omit() # safe there is one row with no data
free_flight[, treatment := ifelse(treatment == "aa", "AA", "CN")]
free_flight[, treatment := factor(treatment, levels = c("CN", "AA"))]
free_flight[, sex := ifelse(sex == "m", "M", "F")]
free_flight[, sex := factor(sex, levels = c("M", "F"))]

tunnel <- free_flight[experiment == "tunnel"]
```

```{r}
freq_f <- lm(f ~ treatment * v1, data = tunnel[sex == "F"])
coef(summary(freq_f))

gg_freq_f <- ggplot(data = tunnel[sex == "F"],
                    aes(x = v1,
                        y = f,
                        color = treatment)) +
  geom_point() +
  geom_ancova(freq_f) +
  theme_pubr()

freq_m <- lm(f ~ treatment * v1, data = tunnel[sex == "M"])
coef(summary(freq_m))

gg_freq_m <- ggplot(data = tunnel[sex == "M"],
                    aes(x = v1,
                        y = f,
                        color = treatment)) +
  geom_point() +
  geom_ancova(freq_m) +
  theme_pubr()

plot_grid(gg_freq_m, gg_freq_f, ncol = 2)
```


```{r free-flight-import}
subfolder <- "kinematics/free flight/processed"
filename <- "init_data.xlsx"
in_path <- here(data_folder, subfolder, filename)

free_flight_init <- read_xlsx(in_path,
                   sheet = "init_data") |>
  clean_names() |>
  data.table()
setnames(free_flight_init, old = "treat", new = "treatment")
free_flight_init[, treatment := ifelse(treatment == "aa", "AA", "CN")]
free_flight_init[, treatment := factor(treatment, levels = c("CN", "AA"))]
free_flight_init[, sex := ifelse(sex == "m", "M", "F")]
free_flight_init[, sex := factor(sex, levels = c("M", "F"))]

free_flight_init[, v_bar := (v1+v2)/2]
```

```{r}
lm1 <- lm(f ~ treatment + v_bar, data = free_flight_init[sex == "F"])
coef(summary(lm1))

ggplot(data = free_flight_init[sex == "F"],
       aes(x = v_bar,
           y = f,
           color = treatment)) +
  geom_point() +
#  geom_ancova(lm1) +
  theme_pubr()
```

```{r}

```


# Wingshape analysis
