---
title: "Drosophila Burst Speed"
author: "Jeff Walker"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, echo=FALSE}
# wrangling packages
library(data.table) # magical data frames
library(here) # here makes a project transportable
library(janitor) # clean_names
library(readxl) # read_excel, duh!
library(writexl) # write_xlsx, duh!
library(stringr) # text wrangling

# analysis packages
library(lme4)
library(lmerTest)
library(emmeans) # the workhorse for inference
library(sf)

# graphing packages
library(ggsci) # color palettes
library(ggpubr) # publication quality plots
library(ggforce) # better jitter
library(cowplot) # combine plots
library(knitr) # kable tables
library(kableExtra) # kable_styling tables
library(jpeg)

# ggplot_the_model.R packages not loaded above
library(insight)
library(afex) # aov4

# use here from the here package
here <- here::here
# use clean_names from the janitor package
clean_names <- janitor::clean_names
# use transpose from data.table
transpose <- data.table::transpose

# load functions used by this text written by me
# ggplot_the_model.R needs to be in the folder "R"
# if you didn't download this and add to your R folder in your
# project, then this line will cause an error

source(here("code", "ggplot_the_model.R"))
source(here("code", "ggptm.R"))
source(here("code", "superimposition.R"))

data_folder <- "data"

# constants
dial_to_speed <- 6.42237212 # from my smoke test with screens and forcing through origin

```

# Burst analysis
## Experiment 1 (Simons-Walker)
### Import
The major analyis is an estimate of the effects.

```{r import}
subfolder <- "burst speed/simons-walker "
filename <- "Wind Tunnel minus days>6.xlsx"
in_path <- here(data_folder, subfolder, filename)

exp1 <- read_xlsx(in_path,
                    range = "A1:H108") |>
  clean_names() |>
  data.table()

exp1[, id := .I]
exp1[, treatment := factor(treatment, levels = c("CN", "AA"))]
exp1[, sex := factor(sex, levels = c("M", "F"))]
exp1[, line := paste0(treatment, line) |>
       factor(levels = c("CN1", "CN2", "AA1", "AA2"))]
exp1[, wind_speed := dial_50 * dial_to_speed]
exp1[, group := ifelse(treatment == "AA", "Selection", "No Selection") |> factor(levels=c("No Selection", "Selection"))]
```

```{r exp1-checks}
exp1[, .(max = max(wind_speed)), by = .(line, sex)]
```

```{r exp1-init-plot}
ggplot(data = exp1,
       aes(x = line,
           y = wind_speed)) +
  geom_jitter(width = 0.1)
```


```{r exp1-sample-means}

exp1_summary <- exp1[, .(wind_speed = mean(wind_speed)), by = .(treatment, sex, line)]
exp1_summary
```

```{r exp1-room-temp}
lm_ancova <- lm(wind_speed ~ treatment * sex + room_temp, data = exp1)
coef(summary(lm_ancova))

ggplot(data = exp1,
       aes(x = room_temp,
           y = wind_speed,
           color = treatment)) +
  geom_point() +
#  geom_ancova(lm_ancova) +
  NULL
```

```{r exp1-balance}
exp1[, .(N = .N), by = .(line, sex)]
```
### Estimate Treatment Effects

```{r exp1-lmm-random-intercept}
exp1_lmm <- lmer(wind_speed ~ treatment * sex + (1 | line), data = exp1)
coef(summary(exp1_lmm))
exp1_sigma <- summary(exp1_lmm)$sigma

exp1_emm <- emmeans(exp1_lmm, specs = c("treatment", "sex"))

cn_m <- c(1,0,0,0)
aa_m <- c(0,1,0,0)
cn_f <- c(0,0,1,0)
aa_f <- c(0,0,0,1)
exp1_contrast <- list(
  list("F: AA - CN" = aa_f - cn_f),
  list("M: AA - CN" = aa_m - cn_m)
)

exp1_pairs <- contrast(exp1_emm,
                      method = exp1_contrast,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
exp1_pairs[, effect_std := estimate / exp1_sigma]
# reorder for plot
exp1_pairs[, contrast := factor(contrast,
                                levels = c("M: AA - CN",
                                           "F: AA - CN"))]

exp1_emm
exp1_pairs

```

### Plot

1. Separate sex
2. plot each line as a separate X value, show the data
3. get line means from lmm

```{r exp1-conditional-means}
# compute conditional means
newdat <- expand.grid(
  line = levels(exp1$line),
  sex = levels(exp1$sex)
) |>
  data.table()
newdat[, treatment := ifelse(substr(newdat$line, 1, 2) == "AA", "AA", "CN")]
newdat[, fit := predict(exp1_lmm, newdata = newdat)]

# check with my own math
coefs <- coef(exp1_lmm)$line
mm_males <- matrix(c(1,0,0,0,
                       1,0,0,0,
                       1,1,0,0,
                       1,1,0,0), nrow=4)
mm_females <- matrix(c(1,0,1,0,
                       1,0,1,0,
                       1,1,1,1,
                       1,1,1,1), nrow=4)
male_means <- as.matrix(coefs) %*% mm_males |> diag()
female_means <- as.matrix(coefs) %*% mm_females |> diag()

# compare with raw means
fixed_means <- exp1[, .(speed = mean(wind_speed)),
                      by = .(sex, line, treatment)]
fixed_means

```

```{r exp1-conditional-mean-plot}

means_plot <- function(plot_sex = "F", title = "Female"){
  offset <- 0.15
  x_breaks <- c(1-offset, 1+offset, 2-offset, 2+offset)
  
  plot_data <- exp1[, .SD, .SDcols = c("treatment", "line", "sex", "wind_speed")]
  treatment_means <- exp1_emm |> 
    summary() |>
    data.table()
  treatment_means[treatment == "CN", x := 1]
  treatment_means[treatment == "AA", x := 2]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "CN1", x := 1 - offset]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "AA1", x := 2 - offset]
  plot_data[line == "AA2", x := 2 + offset]
  newdat[line == "CN1", x := 1 - offset]
  newdat[line == "CN2", x := 1 + offset]
  newdat[line == "AA1", x := 2 - offset]
  newdat[line == "AA2", x := 2 + offset]
  
  gg <- ggplot(data = plot_data[sex == plot_sex],
               aes(x = x,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    geom_jitter(width = 0.05,
                height = 0,
                color = "gray") +
    
    # line means
    geom_point(data = newdat[sex == plot_sex],
               aes(x = x,
                   y = fit),
               size = 2) +
    
    # treatment means and CI
    geom_point(data = treatment_means[sex == plot_sex],
               aes(x = x,
                   y = emmean),
               size = 3) +
    geom_errorbar(data = treatment_means[sex == plot_sex],
                  aes(x = x,
                      y = emmean,
                      ymin = lower.CL,
                      ymax = upper.CL),
                  width = 0) +
    
    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = pal_okabe_ito) +
    scale_x_continuous(breaks = x_breaks,
                       labels = c("CN1", "CN2", "AA1", "AA2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(90, 210),
                    xlim = c(0.7, 2.3)) +
    ggtitle(title)
  return(gg)
}

gg_f <- means_plot(plot_sex = "F", title = "Female")
gg_m <- means_plot(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```

```{r exp1-effect-plot}
gg <- ggplot(data = exp1_pairs,
             aes(y = contrast,
                 x = estimate)) +
  geom_point() +
  geom_errorbar(aes(xmin = lower.CL,
                    xmax = upper.CL),
                width = 0.2) +
  coord_cartesian(xlim = c(-20, 100)) +
  xlab("Effect (cm/s)") +
  ylab("") +
  scale_x_continuous(
    sec.axis = sec_axis(trans = ~ . / exp1_sigma, name = "Standardized Effect")) +
  theme(axis.title.y = element_blank()) +
  theme_pubr()
gg
```

### Other models
With room_temp covariate (supplement)

```{r exp1-lmm-random-intercept}
exp1_lmm_2 <- lmer(wind_speed ~ treatment * sex + room_temp + (1 | line), data = exp1)
coef(summary(exp1_lmm_2))
sigma <- summary(exp1_lmm_2)$sigma

exp1_lmm_2_emm <- emmeans(exp1_lmm_2, specs = c("treatment", "sex"))

exp1_lmm_2_pairs <- contrast(exp1_lmm_2_emm,
                      method = exp1_contrast,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
exp1_lmm_2_pairs[, effect_std := estimate / sigma]

exp1_lmm_2_emm
exp1_lmm_2_pairs

```
```{r exp1-classical-nested-anova}
str(exp1)
m1 <- aov(wind_speed ~ treatment * sex + Error(line / sex),
          data = exp1)
summary(m1)

m1_male <- aov(wind_speed ~ treatment + Error(line),
          data = exp1)
```


## Experiment 2 (Wheeler) data
### Clean and Import
```{r exp2-clean-his-files}
subfolder <- "burst speed/jess wheeler/sent from jess/raw"
filename <- "02.14.10.xlsx"
in_path <- here(data_folder, subfolder, filename)

  file_i <- read_xlsx(in_path,
                    range = "a1:c1000") |>
    clean_names() |>
    data.table() |>
    na.omit()
  # angle is flight angle in degrees from 1,0
  # length is flight distance over 1 frame
  colnames(file_i)[1] <- "id"
  file_i[, treatment := substr(id,1,2)]
  file_i[, line := substr(id,1,3)]
  file_i[, sex := substr(id, 4,4)]
  file_i[, dial := substr(id, 5,6) |> as.integer()]
  file_i[, date := substr(id, 8,15)]
 
  check_n <- file_i[, .(N = .N),
     by = .(date, sex, treatment, line, id)]
  check_n[N >=6, id]

  check_id <- check_n[, .(N = .N), 
     by = .(date, sex, treatment, line)]
setorder(check_id, sex, treatment, line, date)

```

```{r exp2-import-cleaned}

subfolder <- "burst speed/jess wheeler/sent from jess/raw"
subfolder_path <- here(data_folder, subfolder)

files <- list.files(
  path = subfolder_path,
  pattern = "\\.xlsx$",
  full.names = TRUE
)

exp2 <- data.table(NULL)

for(i in 1:length(files)){
  in_path <- files[i]
  
  file_i <- read_xlsx(in_path,
                    range = "a1:c1000") |>
    clean_names() |>
    data.table() |>
    na.omit()
  # angle is flight angle in degrees from 1,0
  # length is flight distance over 1 frame
  colnames(file_i)[1] <- "id"
  file_i[, treatment := substr(id,1,2)]
  file_i[, line := substr(id,1,3) |> 
           toupper() |>
           factor()]
  file_i[, sex := substr(id, 4,4) |> 
           toupper() |>
           factor()]
  file_i[, dial := substr(id, 5,6) |> as.integer()]
  file_i[, date := substr(id, 8,15)]
  
  # performance measures
  file_i[, theta := ifelse(angle < 0,
                         -(angle + 180),
                         180-angle)]
  file_i[, wind_jess := 7.3888 * dial]
  file_i[, wind_jaw := 6.42237212 * dial]
  file_i[, disp.x := length * cos(theta * pi/180)]
  file_i[, disp.y := length * sin(theta * pi/180)]
  file_i[, wind.x := wind_jess / 30]
  file_i[, vel := sqrt((wind.x + disp.x)^2 + disp.y^2) * 30]
  
  exp2 <- rbind(exp2, file_i)
  
}

  # clean treatments and make factor
  exp2[treatment == "aa", treatment := "AA"]
  exp2[treatment == "cn", treatment := "CN"]
  # exp2[treatment == "k4", treatment := "K45"]
  # exp2[treatment == "n6", treatment := "N60"]
  exp2[treatment == "k4", treatment := "Wild"]
  exp2[treatment == "n6", treatment := "Wild"]
  exp2[, treatment := factor(treatment, levels = 
                                 c("CN", "Wild", "AA"))]
  exp2[, line := factor(line, 
                        levels = c("CN1", "CN2",
                                   "K45", "N60",
                                   "AA1", "AA2"))]
  exp2[treatment == "AA", group := "Selection"]
  exp2[treatment == "CN", group := "No Selection"]
  exp2[treatment == "Wild", group := "Wild"]
  exp2[, group := factor(group,
                         levels = c("No Selection",
                                    "Selection",
                                    "Wild"))]

```

```{r exp2-summary}
# this is just to compare counts with Jess' thesis table
# first compute max within each id
exp2_max_id <- exp2[, .(wind_max = mean(wind_jess),
                        vel_max = mean(vel),
                        N = .N),
     by = .(date, treatment, line, sex, id)]

exp2_sum <- exp2_max_id[, .(wind_max = mean(wind_max),
                            vel_max = mean(vel_max),
                    N = .N), 
     by = .(date, sex, treatment, line)]
setorder(exp2_sum, sex, treatment, line, date)
```

### Primary data set

To compare to Exp1 (Simons-Walker) results, I am using wind tunnel speed and not the computed speed of the fly. But I will analyze that in a supplement. Jess has 3 measures per fly; since these are at same tunnel speed, I just need to obtain the summary data as the primary data set.

```{r jess-primary-data}
exp2_primary <- exp2[, .(wind_speed = mean(wind_jess)), 
                     by = .(id, treatment, line, sex, group)]
```

### Estimate Treatment Effects

```{r exp2-lmm-random-intercept}
exp2_lmm <- lmer(wind_speed ~ treatment * sex + (1 | line), data = exp2_primary)
coef(summary(exp2_lmm))
exp2_sigma <- summary(exp2_lmm)$sigma

exp2_emm <- emmeans(exp2_lmm, specs = c("treatment", "sex"))

cn_f <- c(1,0,0,0,0,0)
wild_f <- c(0,1,0,0,0,0)
aa_f <- c(0,0,1,0,0,0)
cn_m <- c(0,0,0,1,0,0)
wild_m <- c(0,0,0,0,1,0)
aa_m <- c(0,0,0,0,0,1)

# contrasts of interest - this is all contrasts, just not m v f
# aa - cn # primary contrast
# aa - k45 # secondary contrast testing comparison with newer line
# aa - n60 # secondary contrast testing comparison with newer line
# cn - k45 # secondary contrast testing loss in cn
# cn - n60 # secondary contrast testing loss in cn
# n60 - k45 # n60 is july 2009 and k45 is july 2008

exp2_contrasts <- list(
  list("F: AA - CN" = aa_f - cn_f),
  list("F: AA - WT" = aa_f - wild_f),
  list("F: CN - WT" = cn_f - wild_f),
  list("M: AA - CN" = aa_m - cn_m),
  list("M: AA - WT" = aa_m - wild_m),
  list("M: CN - WT" = cn_m - wild_m)
)

exp2_pairs <- contrast(exp2_emm,
                      method = exp2_contrasts,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
exp2_pairs[, effect_std := estimate / sigma]
# set up order for effects plot
exp2_pairs[, contrast := factor(contrast,
                                levels = c("M: CN - WT",
                                           "M: AA - WT",
                                           "M: AA - CN",
                                           "F: CN - WT",
                                           "F: AA - WT",
                                           "F: AA - CN"))]

exp2_emm
exp2_pairs

```

### Plot

```{r exp2-conditional-means}
# compute conditional means
newdat <- expand.grid(
  line = levels(exp2_primary$line),
  sex = levels(exp2_primary$sex)
) |>
  data.table()
# newdat[, treatment := line]
# newdat[substr(line,1,2) == "aa", treatment := "aa"]
# newdat[substr(line,1,2) == "cn", treatment := "cn"]
newdat[, treatment := substr(line,1,2)]
newdat[treatment == "K4", treatment := "Wild"]
newdat[treatment == "N6", treatment := "Wild"]
newdat[, fit := predict(exp2_lmm, newdata = newdat)]

```

```{r exp2-conditional-mean-plot}

means_plot <- function(plot_sex = "F", title = "Female"){
  offset <- 0.25
  x_breaks <- c(1-offset, 1+offset, 2-offset, 2+offset,
                3-offset, 3+offset)
  
  plot_data <- exp2_primary[, .SD, .SDcols = c("treatment", "line", "sex", "wind_speed")]
  treatment_means <- exp2_emm |> 
    summary() |>
    data.table()
  treatment_means[treatment == "CN", x := 1]
  treatment_means[treatment == "Wild", x := 2]
  treatment_means[treatment == "AA", x := 3]
  plot_data[line == "CN1", x := 1 - offset]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "K45", x := 2 - offset]
  plot_data[line == "N60", x := 2 + offset]
  plot_data[line == "AA1", x := 3 - offset]
  plot_data[line == "AA2", x := 3 + offset]
  newdat[line == "CN1", x := 1 - offset]
  newdat[line == "CN2", x := 1 + offset]
  newdat[line == "K45", x := 2 - offset]
  newdat[line == "N60", x := 2 + offset]
  newdat[line == "AA1", x := 3 - offset]
  newdat[line == "AA2", x := 3 + offset]

  gg <- ggplot(data = plot_data[sex == plot_sex],
               aes(x = x,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    geom_jitter(width = 0.05,
                height = 0.5,
                size = 0.2,
                color = "gray") +

    # line means
    geom_point(data = newdat[sex == plot_sex],
               aes(x = x,
                   y = fit),
               size = 2) +
    
    # treatment means and CI
    geom_point(data = treatment_means[sex == plot_sex],
               aes(x = x,
                   y = emmean),
               size = 3) +
    geom_errorbar(data = treatment_means[sex == plot_sex],
                  aes(x = x,
                      y = emmean,
                      ymin = lower.CL,
                      ymax = upper.CL),
                  width = 0) +
    
    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = c(pal_okabe_ito[1],
                                  pal_okabe_ito[3],
                                  pal_okabe_ito[2])) +
    scale_x_continuous(breaks = x_breaks,
                       labels = c("CN\n1", "CN\n2",
                                  "Wild\nK45", "Wild\nN60",
                                   "AA\n1", "AA\n2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(0, 250),
                    xlim = c(0.7, 3.3)) +
    ggtitle(title)
  return(gg)
}

gg_f <- means_plot(plot_sex = "F", title = "Female")
gg_m <- means_plot(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```

```{r exp2-effects-plot}
gg <- ggplot(data = exp2_pairs,
             aes(y = contrast,
                 x = estimate)) +
  geom_point() +
  geom_errorbar(aes(xmin = lower.CL,
                    xmax = upper.CL),
                width = 0.2) +
  coord_cartesian(xlim = c(-65, 100)) +
  xlab("Effect (cm/s)") +
  ylab("") +
  scale_x_continuous(
    sec.axis = sec_axis(trans = ~ . / exp1_sigma, name = "Standardized Effect")) +
  theme(axis.title.y = element_blank()) +
  theme_pubr()
gg
```

```{r exp2-conditional-mean-plot-violin}

means_plot <- function(plot_sex = "F", title = "Female"){

  gg <- ggplot(data = exp2_primary[sex == plot_sex],
               aes(x = line,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    # geom_jitter(width = 0.05,
    #             height = 0,
    #             color = "gray") +

    # geom_violin() +
    
    geom_boxplot() +

    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = 
                         c(pal_okabe_ito[1], 
                           pal_okabe_ito[3], 
                           pal_okabe_ito[2])) +
    scale_x_discrete(labels = c("CN\n1", "CN\n2",
                                "Wild\nK45", "Wild\nN60",
                                "AA\n1", "AA\n2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(25, 225)) +
    ggtitle(title)
  return(gg)
}

gg_f <- means_plot(plot_sex = "F", title = "Female")
gg_m <- means_plot(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```


## Combined

```{r burst-combined-plot}
burst1 <- data.table(
  experiment = "Exp1",
  exp1[, .SD, .SDcols = c("treatment", "line", "sex", "group",  "wind_speed")])
burst2 <- data.table(
  experiment = "Exp2",
  exp2[, .SD, .SDcols = c("treatment", "line", "sex", "group", "wind_jess")])
setnames(burst2, old = "wind_jess", new = "wind_speed")
burst <- rbind(burst1, burst2)
burst[, exp_line := paste(experiment, line, sep="\n")]
burst[, exp_line := factor(exp_line, levels = c(
  "Exp1\nCN1", "Exp1\nCN2",
  "Exp1\nAA1", "Exp1\nAA2",
  "Exp2\nCN1", "Exp2\nCN2",
  "Exp2\nAA1", "Exp2\nAA2",
  "Exp2\nK45", "Exp2\nN60"
))]

combined_plot <- function(sex_in = "F", title = "Female"){
  gg <- ggplot(data = burst[sex == sex_in],
               aes(x = exp_line,
                   y = wind_speed,
                   color = group)) +
  geom_jitter(width = 0.2,
              height = 1,
              size = .2) +
  # clean plot
  ylab("Max Opposing Wind Speed (cm/s)") +
  scale_color_manual(values = pal_okabe_ito) +
  theme_pubr() +
  theme(legend.position="none",
        axis.title.x = element_blank()) +
  ggtitle(title)
  return(gg)
}

gg_f <- combined_plot(sex = "F", title = "Female")
gg_f
gg_m <- combined_plot(sex = "M", title = "Male")
gg_m
```

# Kinematic analysis


```{r free-flight-import}
subfolder <- "kinematics/free flight/processed"
filename <- "frequency data.xlsx"
in_path <- here(data_folder, subfolder, filename)

free_flight <- read_xlsx(in_path,
                   sheet = "frequency data") |>
  clean_names() |>
  data.table() |>
  na.omit() # safe there is one row with no data
free_flight[, treatment := ifelse(treatment == "aa", "AA", "CN")]
free_flight[, treatment := factor(treatment, levels = c("CN", "AA"))]
free_flight[, sex := ifelse(sex == "m", "M", "F")]
free_flight[, sex := factor(sex, levels = c("M", "F"))]

tunnel <- free_flight[experiment == "tunnel"]
```

```{r}
freq_f <- lm(f ~ treatment * v1, data = tunnel[sex == "F"])
coef(summary(freq_f))

gg_freq_f <- ggplot(data = tunnel[sex == "F"],
                    aes(x = v1,
                        y = f,
                        color = treatment)) +
  geom_point() +
  geom_ancova(freq_f) +
  theme_pubr()

freq_m <- lm(f ~ treatment * v1, data = tunnel[sex == "M"])
coef(summary(freq_m))

gg_freq_m <- ggplot(data = tunnel[sex == "M"],
                    aes(x = v1,
                        y = f,
                        color = treatment)) +
  geom_point() +
  geom_ancova(freq_m) +
  theme_pubr()

plot_grid(gg_freq_m, gg_freq_f, ncol = 2)
```


```{r free-flight-import}
subfolder <- "kinematics/free flight/processed"
filename <- "init_data.xlsx"
in_path <- here(data_folder, subfolder, filename)

free_flight_init <- read_xlsx(in_path,
                   sheet = "init_data") |>
  clean_names() |>
  data.table()
setnames(free_flight_init, old = "treat", new = "treatment")
free_flight_init[, treatment := ifelse(treatment == "aa", "AA", "CN")]
free_flight_init[, treatment := factor(treatment, levels = c("CN", "AA"))]
free_flight_init[, sex := ifelse(sex == "m", "M", "F")]
free_flight_init[, sex := factor(sex, levels = c("M", "F"))]

free_flight_init[, v_bar := (v1+v2)/2]
```

```{r}
lm1 <- lm(f ~ treatment + v_bar, data = free_flight_init[sex == "F"])
coef(summary(lm1))

ggplot(data = free_flight_init[sex == "F"],
       aes(x = v_bar,
           y = f,
           color = treatment)) +
  geom_point() +
#  geom_ancova(lm1) +
  theme_pubr()
```

```{r}

```


# Wingshape analysis

For wing anatomy, see https://flybase.org/reports/FBim0000853?utm_source=chatgpt.com

and especially
https://pmc.ncbi.nlm.nih.gov/articles/PMC11457947/?utm_source=chatgpt.com

## Aerodynamic wing metrics 

```{r wing-metrics}
wing_metrics <- function(x, y, hinge = 1, tip = 17) {

  ## ------------------------------------------------------------
  ## 1. Translate so hinge (LM1) is at the origin
  ## ------------------------------------------------------------

  x0 <- x - x[hinge]
  y0 <- y - y[hinge]


  ## ------------------------------------------------------------
  ## 2. Rotate so hinge -> tip (LM17) is the radial x-axis
  ## ------------------------------------------------------------

  theta <- atan2(y0[tip], x0[tip])

  xr <-  x0 * cos(theta) + y0 * sin(theta)
  yr <- -x0 * sin(theta) + y0 * cos(theta)

  ## Wing length: hinge -> tip
  R <- sqrt(x0[tip]^2 + y0[tip]^2)


  ## ------------------------------------------------------------
  ## 3. Polygon integrals
  ## ------------------------------------------------------------

  # next vertex around perimeter
  x2 <- c(xr[-1], xr[1])
  y2 <- c(yr[-1], yr[1])

  cross <- xr * y2 - x2 * yr

  # signed area
  As <- sum(cross) / 2

  # make formulas independent of clockwise/counterclockwise digitizing
  s <- sign(As)
  A <- abs(As)

  # centroid
  xbar <- sum((xr + x2) * cross) / (6 * As)
  ybar <- sum((yr + y2) * cross) / (6 * As)

  # second moments of AREA about hinge-defined axes
  #
  # Ixx = integral y^2 dA   (chordwise distribution)
  # Iyy = integral x^2 dA   (radial distribution)
  #
  Ixx <- s * sum(
    (yr^2 + yr*y2 + y2^2) * cross
  ) / 12

  Iyy <- s * sum(
    (xr^2 + xr*x2 + x2^2) * cross
  ) / 12

  # product moment
  Ixy <- s * sum(
    (2*xr*yr + xr*y2 + x2*yr + 2*x2*y2) * cross
  ) / 24


  ## ------------------------------------------------------------
  ## 4. Derived wing metrics
  ## ------------------------------------------------------------

  # Ray-style / single-wing aspect ratio convention
  AR <- R^2 / A

  # mean chord
  mean_chord <- A / R

  # radial position of area centroid
  xbar_hat <- xbar / R

  # radius of second moment of area
  r2 <- sqrt(Iyy / A)

  # nondimensional radius of second moment
  r2_hat <- r2 / R
  
  # nondimensional second moment of area
  r2_hat_sq <- Iyy / (A * R^2)


  ## ------------------------------------------------------------
  ## 5. Return results
  ## ------------------------------------------------------------

  data.table(
    area       = A,
    wing_length = R,
    mean_chord = mean_chord,
    aspect_ratio = AR,

    xbar       = xbar,
    ybar       = ybar,
    xbar_hat   = xbar_hat,

    Ixx        = Ixx,
    Iyy        = Iyy,
    Ixy        = Ixy,

    r2         = r2,
    r2_hat     = r2_hat,
    r2_hat_sq  = r2_hat_sq
  )
}
```

## Import wing shape data
The list of folders are in data/wing shape/wing_files_list.txt

```{r import-wing-data}
list_in <- "data/wing shape/wing_files_list.txt"
list_path <- here(list_in)
folder_list <- fread(list_path, sep = "\t")[[1]]

aero_labels <- c("area", "wing_length","mean_chord", "aspect_ratio", "xbar", "ybar", "xbar_hat", "Ixx", "Iyy", "Ixy", "r2", "r2_hat", "r2_hat_sq")
radii <- c(.1, 0.2, 0.3)
area_labels <- paste0("q", radii)
area_prop_labels <- c("prop_prox")
measure_labels <- c(aero_labels, area_labels, area_prop_labels)

wing_lm <- data.table(NULL) # wing landmarks
wing <- data.table(NULL) # wing metrics
for(i in 1:length(folder_list)){
  folder_i <- folder_list[i]
  rel_folder_path <- paste0("data/wing shape/",folder_i)
  folder_path <- here(rel_folder_path)
  files_short <- list.files( # just the file name
    path = folder_path,
    pattern = "\\.txt$",
    full.names = FALSE
  )
  files_long <- list.files( # full address
    path = folder_path,
    pattern = "\\.txt$",
    full.names = TRUE
  )
  n_files <- length(files_long)
  n_measures <- length(measure_labels)
  wing_measures <- matrix(NA, nrow = n_files, ncol = n_measures)
  colnames(wing_measures) <- measure_labels
  fig_mat <- matrix(NA, nrow = n_files, ncol = 37 * 2)
  colnames(fig_mat) <- c(paste0("x", 1:37),
                         paste0("y", 1:37))
  valid <- rep(FALSE, n_files)
  for(j in 1:n_files){
    lines <- readLines(files_long[j], warn = FALSE)
    lines <- lines[nzchar(trimws(lines))]
    n_landmarks <- length(lines) - 1
    
    if(n_landmarks != 37){
      cat(files_short[j], ":", n_landmarks, "landmarks\n")
      next
    }
    
    valid[j] <- TRUE
    
    fig_lm <- fread(files_long[j], nrows = 37)
    fig_lm[, V2 := -V2]
    
    # save lm to a matrix for the all fly_id in folder
    fig_mat[j,] <- c(unlist(fig_lm[, V1]),
                     unlist(fig_lm[, V2]))
    # compute wing metrics and save for all fly_id in folder
    aero_j <- wing_metrics(x = fig_lm[, V1],
                           y = fig_lm[, V2])
    area_j <- radial_area_distribution(x = fig_lm[, V1],
                                       y = fig_lm[, V2],
                                       q = radii)
    regional_j <- regional_areas(
      x = fig_lm[, V1],
      y = fig_lm[, V2]
    )
    combined_j <- c(unlist(aero_j[1,]),
                    unlist(area_j[,2]),
                    unlist(regional_j[1, "prop_prox"])
    )
    wing_measures[j, ] <- combined_j
    
  }
  # delete empty rows
  # wing_measures <- na.omit(wing_measures)
  # fig_mat <- na.omit(fig_mat)
  files_short   <- files_short[valid]
  fig_mat       <- fig_mat[valid, , drop = FALSE]
  wing_measures <- wing_measures[valid, , drop = FALSE]
  
  # save landmarks
  wing_lm <- rbind(
    wing_lm,
    data.table(
      file = str_remove(files_short, "\\.txt$"),
      temp_id = str_sub(files_short, 2, 2),
      line_id = str_sub(files_short, 3, 4),
      sex = str_sub(files_short, 5, 5),
      exp_id = str_sub(files_short, 6, 6),
      fly_id = str_extract(files_short, "(?<=^.{6})\\d{1,2}"),
      fig_mat
    )
  )
  # save wing metrics
  wing <- rbind(
    wing,
    data.table(
      file = str_remove(files_short, "\\.txt$"),
      temp_id = str_sub(files_short, 2, 2),
      line_id = str_sub(files_short, 3, 4),
      sex = str_sub(files_short, 5, 5),
      exp_id = str_sub(files_short, 6, 6),
      fly_id = str_extract(files_short, "(?<=^.{6})\\d{1,2}"),
      wing_measures
    )
  )
}

wing[, r2_hat_sq2 := r2_hat^2] # see Ray et al 2016

wing[, treatment := ifelse(line_id == "A1" |
                             line_id == "A2",
                           "SEL", "WT") |>
       factor(levels = c("WT", "SEL"))]

wing_lm[, treatment := ifelse(line_id == "A1" |
                                line_id == "A2",
                              "SEL", "WT") |>
          factor(levels = c("WT", "SEL"))]
```


## Analysis
### Plot superimposed wings by treatment combination

```{r plot-superimposed-wings}
# get tps coords
  all_cols <- 1:37
  lm_subset <- c(1,3,5,12,17,21,27,37)
  keep_cols <- all_cols
  p <- length(keep_cols)

  lm_cols <- do.call(paste0,
                     expand.grid(c("x","y"), keep_cols))
  lm_mat <- wing_lm[, .SD, .SDcols = lm_cols] |>
    as.matrix()
  lm_array <- matrix_2_array(lm_mat)
  lm_tpr <- tpr(lm_array, 1, 17)

  lm_dt_tpr <- array_2_matrix(lm_tpr) |>
    data.table()
  lm_cols.tpr <- paste0(lm_cols, ".tpr")
  colnames(lm_dt_tpr) <- lm_cols.tpr
  wing_tpr <- cbind(wing_lm, lm_dt_tpr)
  
  lm_means <- wing_tpr[, lapply(.SD, mean),
                         by = .(treatment),
                         .SDcols = c(lm_cols.tpr)]
  lm_mat <- lm_means[, .SD, .SDcols = lm_cols.tpr] |>
    as.matrix()
  lm_tpr <- matrix_2_array(lm_mat)
  gg <- plot_2shapes(lm_tpr, 
                     color_col = rep(c("SEL", "WT"), each = p))
  gg
  

```

```{r plot-superimposed-wings-by-treatment-combi}
# treatment x temp x experiment x sex
# 2 x 3 x 2 x 2
lm_means <- wing_tpr[, lapply(.SD, mean),
                    by = .(treatment, temp_id,
                           exp_id, sex),
                    .SDcols = c(lm_cols.tpr)]

design <- CJ(
#  sex     = c("F", "M"),
  sex     = c("F"),
  temp_id = c("X", "Y", "Z"),
  exp_id  = c("A", "B")
)
plots <- vector("list", nrow(design))
for(i in 1:nrow(design)){
  lm_ss <- lm_means[temp_id == design[i, temp_id] &
                      exp_id == design[i, exp_id] &
                      sex == design[i, sex],
                    .SD, .SDcols = c("treatment", lm_cols.tpr)]
  lm_mat <- lm_ss[, .SD, .SDcols = lm_cols.tpr] |>
    as.matrix()
  lm_tpr <- matrix_2_array(lm_mat)
  plots[[i]] <- plot_2shapes(lm_tpr, 
                      color_col = rep(c("SEL", "WT"), each = p)) +
    ggtitle(paste(design[i, temp_id],
                  design[i, exp_id],
                  design[i, sex])) +
    theme(legend.position = "none")
}

gg <- plot_grid(plotlist = plots,
          nrow = 3)

ggsave(
  "wing_shapes.png",
  gg,
  width = 8,
  height = 12,
  dpi = 300
)

gg
```

### Reconstruct landmarks on image to identify landmarks

Don't need this because I have the image from Jim

```{r reconstruct-landmarks-on-wing-image}

# run this in console
# quartz()
# then run the code and click on the 4 landmarks where wing veins intersect the perimeter
image_file <- "wing_image/WXA1FA2.jpg"
image_file <- "wing_image/WXA2FA1.jpg"
image_path <- here(image_file)
img <- readJPEG(image_path)


h <- dim(img)[1]
w <- dim(img)[2]

plot(
  NA,
  xlim = c(0, w),
  ylim = c(h, 0),
  asp = 1,
  xaxs = "i",
  yaxs = "i",
  xlab = "",
  ylab = ""
)

rasterImage(img, 0, h, w, 0)


known <- c(12, 17, 21, 27)
# Display image first, then click landmarks in exactly this order:
# 12, 17, 21, 27
pix <- locator(4)

pixel <- data.table(
  point = known,
  X = pix$x,
  Y = pix$y
)
```

```{r reconstruct-part-ii}
# registration coordinates
reg_file <- "wing_image/WXA2FA1_reg.txt"
reg_path <- here(reg_file)
pixel <- fread(reg_path)

# landmarks
lm_file <- "wing_image/WXA2FA1pp.txt"
lm_path <- here(lm_file)
xy <- fread(lm_path, nrows = 37,
            col.names = c("x", "y"))

xy[, point := 1:.N]

reg <- merge(xy, pixel, by = "point")

M <- rbind(
  cbind(1, 0,  reg$x, -reg$y),
  cbind(0, 1,  reg$y,  reg$x)
)

z <- c(reg$X, reg$Y)

A <- do.call(rbind, lapply(seq_len(nrow(reg)), function(i) {
  x <- reg$x[i]
  y <- reg$y[i]

  rbind(
    c(1, 0, x, -y),
    c(0, 1, y,  x)
  )
}))

z <- as.vector(t(as.matrix(reg[, .(X, Y)])))

coef <- qr.solve(A, z)

a <- coef[1]
d <- coef[2]
b <- coef[3]
c <- coef[4]

# transform all points
xy[, X := a + b*x - c*y]
xy[, Y := d + c*x + b*y]

# ggplot
ggplot() +
  annotation_raster(
    img,
    xmin = 0,
    xmax = w,
    ymin = h,
    ymax = 0
  ) +
  geom_path(
    data = xy,
    aes(X, Y),
    linewidth = 0.4,
    color = "red"
  ) +
  geom_point(
    data = xy,
    aes(X, Y),
    color = "red",
    size = 2
  ) +
  geom_text(
    data = xy,
    aes(X, Y, label = point),
    nudge_y = -15,
    size = 3
  ) +
  scale_y_reverse(limits = c(h, 0)) +
  coord_fixed(
    xlim = c(0, w),
    ylim = c(h, 0),
    expand = FALSE
  ) +
  theme_void()
```



### Checking apparent disconnect between Harmon's polygons and my 2nd moment of area

```{r radial-area-distribution}
radial_area_distribution <- function(x, y, hinge = 1, tip = 17,
                                     q = seq(0, 1, by = 0.05)) {

  ## Translate hinge to origin
  x0 <- x - x[hinge]
  y0 <- y - y[hinge]

  ## Rotate hinge -> tip onto x-axis
  theta <- atan2(y0[tip], x0[tip])

  xr <-  x0 * cos(theta) + y0 * sin(theta)
  yr <- -x0 * sin(theta) + y0 * cos(theta)

  ## Wing length
  R <- sqrt(x0[tip]^2 + y0[tip]^2)

  ## Close polygon for sf
  xy <- cbind(xr, yr)
  xy <- rbind(xy, xy[1, ])

  wing <- st_polygon(list(xy))

  total_area <- as.numeric(st_area(wing))

  ## Bounds safely beyond wing
  ymin <- min(yr) - R
  ymax <- max(yr) + R
  xmin <- min(xr) - R

  ## For each q, calculate area with x/R <= q
  cumulative_area <- sapply(q, function(qi) {

    xmax <- qi * R

    box <- st_polygon(list(matrix(
      c(xmin, ymin,
        xmax, ymin,
        xmax, ymax,
        xmin, ymax,
        xmin, ymin),
      ncol = 2,
      byrow = TRUE
    )))

    intersection <- suppressWarnings(
      st_intersection(wing, box)
    )

    if (length(intersection) == 0) {
      0
    } else {
      as.numeric(st_area(intersection)) / total_area
    }
  })

  data.table(
    q = q,
    cumulative_area = cumulative_area
  )
}
```

```{r check-daves-regional-area}
polygon_area <- function(x, y) {
  
  # next point around polygon, including last -> first
  x2 <- c(x[-1], x[1])
  y2 <- c(y[-1], y[1])
  
  abs(sum(x * y2 - x2 * y) / 2)
}


regional_areas <- function(x, y) {
  
  # Total wing
  total <- polygon_area(x, y)
  
  # Proximal: 1 -> 2 -> 3 -> 4 -> 5 -> 37 -> 1
  prox_ind <- c(1:5, 37)
  proximal <- polygon_area(x[prox_ind], y[prox_ind])
  
  # Middle: 5 -> ... -> 12 -> 27 -> ... -> 37 -> 5
  mid_ind <- c(5:12, 27:37)
  middle <- polygon_area(x[mid_ind], y[mid_ind])
  
  # Distal: 12 -> ... -> 27 -> 12
  dist_ind <- 12:27
  distal <- polygon_area(x[dist_ind], y[dist_ind])
  
  data.table(
    area_total = total,
    area_prox = proximal,
    area_middle = middle,
    area_distal = distal,
    
    prop_prox = proximal / total,
    prop_middle = middle / total,
    prop_distal = distal / total,
    
    partition_error =
      (proximal + middle + distal) - total
  )
}
```

### Effects

```{r simple-model}
m1f <- lm(area ~ treatment * sex, data = wing)
m1f_emm <- emmeans(m1f, specs = c("treatment", "sex"))
m1f_pairs <- contrast(m1f_emm,
                      method = "revpairwise",
                      adjust = "none") |>
  summary(infer = TRUE)
ggplot_the_model(m1f, m1f_emm, m1f_pairs)

m1f <- lm(aspect_ratio ~ treatment * sex, data = wing)
m1f_emm <- emmeans(m1f, specs = c("treatment", "sex"))
m1f_pairs <- contrast(m1f_emm,
                      method = "revpairwise",
                      adjust = "none") |>
  summary(infer = TRUE)
ggplot_the_model(m1f, m1f_emm, m1f_pairs)

m1f <- lm(r2_hat_sq ~ treatment * sex, data = wing)
m1f_emm <- emmeans(m1f, specs = c("treatment", "sex"))
m1f_pairs <- contrast(m1f_emm,
                      method = "revpairwise",
                      adjust = "none") |>
  summary(infer = TRUE)
ggplot_the_model(m1f, m1f_emm, m1f_pairs)

```

```{r}
percent_change <- function(y_label){
  r_form <- paste0("log(", y_label,
  ") ~ treatment * temp_id * sex * exp_id + (1 | line_id)")
  m1 <- lmer(r_form, data = wing)
  m1_emm <- emmeans(m1,
                    ~ treatment,
                    weights = "equal"
  )
#  m1_pairs <- contrast(m1_emm, method = "revpairwise")
  m1_pairs <- contrast(m1_emm, method = list("SEL - WT" = c(-1, 1)))
  pairs_dt <- m1_pairs |>
    summary() |>
    data.table()
  estimate <- pairs_dt[1, "estimate"]
  change <- 100 * (exp(estimate) - 1)
  return(unlist(c(estimate, change)))
}

change_matrix <- matrix(NA, nrow = 3, ncol = 2)
colnames(change_matrix) <- c("estimate", "percent")
change_matrix[1, ] <- percent_change("area")
change_matrix[2, ] <- percent_change("aspect_ratio")
change_matrix[3, ] <- percent_change("r2_hat_sq")
change_dt <- data.table(
  measure = c("area", "aspect_ratio", "r2_hat_sq"),
  change_matrix
)




```

```{r}
percent_change_2 <- function(y_label){
  r_form <- paste0("log(", y_label,
  ") ~ treatment * temp_id * sex * exp_id + (1 | line_id)")
  m1 <- lmer(r_form, data = wing)
  m1_emm <- emmeans(m1,
                    ~ treatment | temp_id * sex * exp_id
  )
  m1_pairs <- contrast(m1_emm, method = list("Sel - CN" = c(-1, 1)))
  m1_pairs_df <- as.data.frame(m1_pairs)
  m1_pairs_df$percent <- 100 * (exp(m1_pairs_df$estimate) - 1)
  return(m1_pairs_df)
}

keep_cols <- c("temp_id","sex","exp_id","percent")
res <- percent_change_2("area")
change_dt <- res[, keep_cols]
colnames(change_dt)[4] <- "area"

res <- percent_change_2("wing_length")
change_dt$wing_length <- as.data.frame(res)$percent

res <- percent_change_2("aspect_ratio")
change_dt$aspect_ratio <- as.data.frame(res)$percent

res <- percent_change_2("xbar_hat")
change_dt$xbar_hat <- as.data.frame(res)$percent

res <- percent_change_2("r2_hat_sq")
change_dt$r2_hat_sq <- as.data.frame(res)$percent

change_dt
clipr::write_clip(change_dt)
```

