---
title: "Wind Tunnel Flies Analysis"
author: "Jeffrey Walker"
date: "`r Sys.Date()`"
format: html
editor: source
---

```{r setup, echo=FALSE, message=FALSE}
# wrangling packages
library(data.table) # magical data frames
library(here) # here makes a project transportable
library(janitor) # clean_names
library(readxl) # read_excel, duh!
library(writexl) # write_xlsx, duh!
library(stringr) # text wrangling

# analysis packages
library(lme4)
library(lmerTest)
library(emmeans) # the workhorse for inference
library(sf)

# graphing packages
library(ggsci) # color palettes
library(ggpubr) # publication quality plots
library(ggforce) # better jitter
library(cowplot) # combine plots
library(knitr) # kable tables
library(kableExtra) # kable_styling tables
library(jpeg)

# ggplot_the_model.R packages not loaded above
library(insight)
library(afex) # aov4

# use here from the here package
here <- here::here
# use clean_names from the janitor package
clean_names <- janitor::clean_names
# use transpose from data.table
transpose <- data.table::transpose

# load functions used by this text written by me
# ggplot_the_model.R needs to be in the folder "R"
# if you didn't download this and add to your R folder in your
# project, then this line will cause an error

source(here("code", "ggplot_the_model.R"))
source(here("code", "ggptm.R"))
source(here("code", "superimposition.R"))

data_folder <- "data"

# constants
dial_to_speed <- 6.42237212 # from my smoke test with screens and forcing through origin

```

# Selected lines have continued to increase mean apparent flying speed for 30+ years

Selection started on or before 1991 and continued until 2023.

```{r}
#| label: import-historic-wind-tunnel-data
#| echo: false
#| warning: false
#| eval: true

subfolder <- "wind tunnel"
filename <- "CURRENT WIND TUNNEL GENERATION MEANS.xlsx"
in_path <- here(data_folder, subfolder, filename)

wt_wide <- read_xlsx(in_path,
                sheet = "Sheet1",
                range = "A1:E443") |>
#  clean_names() |>
  data.table()
wt_wide <- wt_wide[gen != ""]
wt <- melt(wt_wide,
           id.vars = "gen",
           measure.vars = c("AA1-1",
                            "AA1-2",
                            "AA2-1",
                            "AA2-2"),
           variable.name = "id",
           value.name = "mafs")
wt[, line := substr(id,1,3)]
wt <- na.omit(wt)

# compute line means (in some generations there were two runs per line)
line_means <- wt[, .(mafs = mean(mafs)),
                 by = .(gen, line)]
aa_mean <- line_means[, .(mafs = mean(mafs)),
                 by = .(gen)]

```

```{r}
#| label: plot-historic-wt-data
#| echo: false
#| warning: false
#| eval: true

gg <- ggplot(data = wt,
             aes(x = gen,
                 y = mafs,
                 color = line)) +
  geom_point(size = 0.5) +
  
  # each AA line is separate
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 20),
    method.args = list(method = "REML"),
    se = FALSE
  ) +
  
  # Combine aa lines
  # geom_smooth(
  #   data = aa_mean,
  #   aes(x = gen, y = mafs),
  #   method = "gam",
  #   formula = y ~ s(x, k = 10),
  #   method.args = list(method = "REML"),
  #   se = FALSE,
  #   inherit.aes = FALSE,
  #   color = "black",
  #   linewidth = 1.2
  # ) +

  
  scale_color_manual(values = pal_okabe_ito) +
  xlab("Generation") +
  ylab("Mean Apparent Flying Speed (cm/s)") +
  theme_pubr()
gg
```

Concerns:

The AA1 and AA2 lines have remarkably similar fit, which could result from

1. Climate (temperature/humidity) fluctuations in the room during selection runs
2. Climate (temperature/humidity/light) fluctions in the rearing room. Since 2001 the temperature and humidity have been well controlled.
3. Husbandry modifications
4. Gene flow between lines

```{r}
#| label: check-increase-in-aa
#| echo: false
#| warning: false
#| eval: false

late150 <- line_means[gen >= 150]
late150[, gen100 := (gen - 150) / 100]

m_150 <- gls(
  mafs ~ gen100 + line,
  data = late150,
  correlation = corCAR1(form = ~ gen | line),
  method = "REML"
)
summary(m_150)
intervals(m_150)

late250 <- line_means[gen >= 250]
late250[, gen100 := (gen - 250) / 100]

m_250 <- gls(
  mafs ~ gen100 + line,
  data = late250,
  correlation = corCAR1(form = ~ gen | line),
  method = "REML"
)
summary(m_250)
intervals(m_250)
```

# Selected lines have 30-40% higher upwind flight ability than control or wild type lines
## Experiment 1 (Simons-Walker data)

```{r }
#| label: import-walker-burst-data
#| echo: false
#| warning: false
#| eval: true

subfolder <- "burst speed/simons-walker "
filename <- "Wind Tunnel minus days>6.xlsx"
in_path <- here(data_folder, subfolder, filename)

exp1 <- read_xlsx(in_path,
                    range = "A1:H108") |>
  clean_names() |>
  data.table()

exp1[, id := .I]
exp1[, treatment := factor(treatment, levels = c("CN", "AA"))]
exp1[, sex := factor(sex, levels = c("F", "M"))]
exp1[, line := paste0(treatment, line) |>
       factor(levels = c("CN1", "CN2", "AA1", "AA2"))]
exp1[, wind_speed := dial_50 * dial_to_speed]
exp1[, group := ifelse(treatment == "AA", "Selection", "No Selection") |> factor(levels=c("No Selection", "Selection"))]
```

```{r exp1-checks}
#| label: exp1-max-check
#| echo: false
#| warning: false
#| eval: false

exp1[, .(max = max(wind_speed)), by = .(line, sex)]
```

```{r}
#| label: exp1-init-plot
#| echo: false
#| warning: false
#| eval: false

ggplot(data = exp1,
       aes(x = line,
           y = wind_speed)) +
  geom_jitter(width = 0.1)
```


```{r }
#| label: exp1-means-check
#| echo: false
#| warning: false
#| eval: false

exp1_summary <- exp1[, .(wind_speed = mean(wind_speed)), by = .(treatment, sex, line)]
exp1_summary
```

### Linear mixed model

```{r}
#| label: exp1-lmm-random-intercept
#| echo: true
#| warning: false
#| eval: true

exp1_lmm <- lmer(wind_speed ~ treatment * sex + (1 | line), data = exp1)
```

```{r }
#| label: exp1-lmm-stats
#| echo: false
#| warning: false
#| eval: true

exp1_sigma <- summary(exp1_lmm)$sigma

exp1_emm <- emmeans(exp1_lmm, specs = c("treatment", "sex"))

cn_f <- c(1,0,0,0)
aa_f <- c(0,1,0,0)
cn_m <- c(0,0,1,0)
aa_m <- c(0,0,0,1)
exp1_contrast <- list(
  list("F: AA - CN" = aa_f - cn_f),
  list("M: AA - CN" = aa_m - cn_m)
)

exp1_pairs <- contrast(exp1_emm,
                      method = exp1_contrast,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
# standardized effect
exp1_pairs[, effect_std := estimate / exp1_sigma]

# percent effect
exp1_cn_means <- summary(exp1_emm)[c(1,3), "emmean"]
exp1_pairs[, effect_perc := estimate / exp1_cn_means * 100]

# reorder for plot
exp1_pairs[, contrast := factor(contrast,
                                levels = c("M: AA - CN",
                                           "F: AA - CN"))]
```


```{r }
#| label: exp1-lmm-emm
#| echo: false
#| warning: false
#| eval: true
exp1_emm |>
  summary() |>
  data.table() |>
  kable(digits = c(1,1,1,2,1,1,1)) |>
  kable_styling()
```


```{r }
#| label: exp1-lmm-pairs
#| echo: false
#| warning: false
#| eval: true
exp1_pairs |>
  kable(digits = c(1,1,2,1,1,1,1,3,1)) |>
  kable_styling()
```


```{r exp1-conditional-means}
#| label: exp1-conditional-means
#| echo: false
#| warning: false
#| eval: true

# compute conditional means
newdat <- expand.grid(
  line = levels(exp1$line),
  sex = levels(exp1$sex)
) |>
  data.table()
newdat[, treatment := ifelse(substr(newdat$line, 1, 2) == "AA", "AA", "CN")]
newdat[, fit := predict(exp1_lmm, newdata = newdat)]

# check with my own math
coefs <- coef(exp1_lmm)$line
mm_females <- matrix(c(1,0,0,0,
                       1,0,0,0,
                       1,1,0,0,
                       1,1,0,0), nrow=4)
mm_males <- matrix(c(1,0,1,0,
                       1,0,1,0,
                       1,1,1,1,
                       1,1,1,1), nrow=4)
male_means <- as.matrix(coefs) %*% mm_males |> diag()
female_means <- as.matrix(coefs) %*% mm_females |> diag()

# compare with raw means
fixed_means <- exp1[, .(speed = mean(wind_speed)),
                      by = .(sex, line, treatment)]
# fixed_means

```

```{r }
#| label: exp1-conditional-mean-plot
#| echo: false
#| warning: false
#| eval: true

means_plot <- function(plot_sex = "F", title = "Female"){
  offset <- 0.15
  x_breaks <- c(1-offset, 1+offset, 2-offset, 2+offset)
  
  plot_data <- exp1[, .SD, .SDcols = c("treatment", "line", "sex", "wind_speed")]
  treatment_means <- exp1_emm |> 
    summary() |>
    data.table()
  treatment_means[treatment == "CN", x := 1]
  treatment_means[treatment == "AA", x := 2]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "CN1", x := 1 - offset]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "AA1", x := 2 - offset]
  plot_data[line == "AA2", x := 2 + offset]
  newdat[line == "CN1", x := 1 - offset]
  newdat[line == "CN2", x := 1 + offset]
  newdat[line == "AA1", x := 2 - offset]
  newdat[line == "AA2", x := 2 + offset]
  
  gg <- ggplot(data = plot_data[sex == plot_sex],
               aes(x = x,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    geom_jitter(width = 0.05,
                height = 0,
                color = "gray") +
    
    # line means
    geom_point(data = newdat[sex == plot_sex],
               aes(x = x,
                   y = fit),
               size = 2) +
    
    # treatment means and CI
    geom_point(data = treatment_means[sex == plot_sex],
               aes(x = x,
                   y = emmean),
               size = 3) +
    geom_errorbar(data = treatment_means[sex == plot_sex],
                  aes(x = x,
                      y = emmean,
                      ymin = lower.CL,
                      ymax = upper.CL),
                  width = 0) +
    
    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = pal_okabe_ito) +
    scale_x_continuous(breaks = x_breaks,
                       labels = c("CN1", "CN2", "AA1", "AA2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(90, 210),
                    xlim = c(0.7, 2.3)) +
    ggtitle(title)
  return(gg)
}

gg_f <- means_plot(plot_sex = "F", title = "Female")
gg_m <- means_plot(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```

### Supplementary models (in supplement)
#### Classical ANCOVA

```{r}
#| label: exp1-lm-room-temp
#| echo: true
#| warning: false
#| eval: true

lm_ancova <- lm(wind_speed ~ treatment * sex + room_temp, data = exp1)

```

```{r }
#| label: exp1-lm-room-temp-stats
#| echo: false
#| warning: false
#| eval: true

coef(summary(lm_ancova))

ggplot(data = exp1,
       aes(x = room_temp,
           y = wind_speed,
           color = treatment)) +
  geom_point() +
#  geom_ancova(lm_ancova) +
  NULL
```

#### Random intercept with covariate

```{r }
#| label: exp1-lmm-room-temp
#| echo: true
#| warning: false
#| eval: true

exp1_lmm_2 <- lmer(wind_speed ~ treatment * sex + room_temp + (1 | line), data = exp1)
```


```{r }
#| label: exp1-lmm-room-temp-stats
#| echo: false
#| warning: false
#| eval: true

sigma <- summary(exp1_lmm_2)$sigma

coef_table <- coef(summary(exp1_lmm_2))
colnames(coef_table)[5] <- "p-value"
coef_dt <- data.table(
  coefficient = row.names(coef_table),
  coef_table
)
coef_table[2:5,] |>
  kable(digits = c(1,2,1,1,3)) |>
  kable_styling()


exp1_lmm_2_emm <- emmeans(exp1_lmm_2, specs = c("treatment", "sex"))

exp1_lmm_2_pairs <- contrast(exp1_lmm_2_emm,
                      method = exp1_contrast,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
exp1_lmm_2_pairs[, effect_std := estimate / sigma]

exp1_lmm_2_pairs|>
  kable(digits = c(1,1,2,1,1,1,1,3,1)) |>
  kable_styling()
```

#### Classical nested ANOVA

```{r }
#| label: exp1-classical-nested-anova
#| echo: true
#| warning: false
#| eval: true

exp1_aov <- aov(wind_speed ~ treatment * sex + Error(line / sex),
          data = exp1)
```


```{r }
#| label: exp1-classical-nested-anova-stats
#| echo: true
#| warning: false
#| eval: true
 
summary(exp1_aov)

```

## Experiment 2 (Wheeler) data
### Clean and Import
```{r }
#| label: exp2-clean-his-files
#| echo: false
#| warning: false
#| eval: false
 
# this chunk is to check and clean all of Jess' excel files, which had duplicate entries
subfolder <- "burst speed/jess wheeler/sent from jess/raw"
filename <- "02.14.10.xlsx"
in_path <- here(data_folder, subfolder, filename)

  file_i <- read_xlsx(in_path,
                    range = "a1:c1000") |>
    clean_names() |>
    data.table() |>
    na.omit()
  # angle is flight angle in degrees from 1,0
  # length is flight distance over 1 frame
  colnames(file_i)[1] <- "id"
  file_i[, treatment := substr(id,1,2)]
  file_i[, line := substr(id,1,3)]
  file_i[, sex := substr(id, 4,4)]
  file_i[, dial := substr(id, 5,6) |> as.integer()]
  file_i[, date := substr(id, 8,15)]
 
  check_n <- file_i[, .(N = .N),
     by = .(date, sex, treatment, line, id)]
  check_n[N >=6, id]

  check_id <- check_n[, .(N = .N), 
     by = .(date, sex, treatment, line)]
setorder(check_id, sex, treatment, line, date)

```

```{r }
#| label: exp2-import-cleaned
#| echo: false
#| warning: false
#| eval: true
 
subfolder <- "burst speed/jess wheeler/sent from jess/raw"
subfolder_path <- here(data_folder, subfolder)

files <- list.files(
  path = subfolder_path,
  pattern = "\\.xlsx$",
  full.names = TRUE
)

exp2 <- data.table(NULL)

for(i in 1:length(files)){
  in_path <- files[i]
  
  file_i <- read_xlsx(in_path,
                    range = "a1:c1000") |>
    clean_names() |>
    data.table() |>
    na.omit()
  # angle is flight angle in degrees from 1,0
  # length is flight distance over 1 frame
  colnames(file_i)[1] <- "id"
  file_i[, treatment := substr(id,1,2)]
  file_i[, line := substr(id,1,3) |> 
           toupper() |>
           factor()]
  file_i[, sex := substr(id, 4,4) |> 
           toupper() |>
           factor()]
  file_i[, dial := substr(id, 5,6) |> as.integer()]
  file_i[, date := substr(id, 8,15)]
  
  # performance measures
  file_i[, theta := ifelse(angle < 0,
                         -(angle + 180),
                         180-angle)]
  file_i[, wind_jess := 7.3888 * dial]
  file_i[, wind_jaw := 6.42237212 * dial]
  file_i[, disp.x := length * cos(theta * pi/180)]
  file_i[, disp.y := length * sin(theta * pi/180)]
  file_i[, wind.x := wind_jess / 30]
  file_i[, vel := sqrt((wind.x + disp.x)^2 + disp.y^2) * 30]
  
  exp2 <- rbind(exp2, file_i)
  
}

  # clean treatments and make factor
  exp2[treatment == "aa", treatment := "AA"]
  exp2[treatment == "cn", treatment := "CN"]
  # exp2[treatment == "k4", treatment := "K45"]
  # exp2[treatment == "n6", treatment := "N60"]
  exp2[treatment == "k4", treatment := "Wild"]
  exp2[treatment == "n6", treatment := "Wild"]
  exp2[, treatment := factor(treatment, levels = 
                                 c("CN", "Wild", "AA"))]
  exp2[, line := factor(line, 
                        levels = c("CN1", "CN2",
                                   "K45", "N60",
                                   "AA1", "AA2"))]
  exp2[treatment == "AA", group := "Selection"]
  exp2[treatment == "CN", group := "No Selection"]
  exp2[treatment == "Wild", group := "Wild"]
  exp2[, group := factor(group,
                         levels = c("No Selection",
                                    "Selection",
                                    "Wild"))]

```

```{r }
#| label: exp2-summary
#| echo: false
#| warning: false
#| eval: false
 
# this is just to compare counts with Jess' thesis table
# first compute max within each id
exp2_max_id <- exp2[, .(wind_max = mean(wind_jess),
                        vel_max = mean(vel),
                        N = .N),
     by = .(date, treatment, line, sex, id)]

exp2_sum <- exp2_max_id[, .(wind_max = mean(wind_max),
                            vel_max = mean(vel_max),
                    N = .N), 
     by = .(date, sex, treatment, line)]
setorder(exp2_sum, sex, treatment, line, date)
```

