---
title: "Wind Tunnel Flies Analysis"
author: "Jeffrey Walker"
date: "`r Sys.Date()`"
format: html
editor: source
---

```{r setup, echo=FALSE, message=FALSE}
# wrangling packages
library(data.table) # magical data frames
library(here) # here makes a project transportable
library(janitor) # clean_names
library(readxl) # read_excel, duh!
library(writexl) # write_xlsx, duh!
library(stringr) # text wrangling

# analysis packages
library(lme4)
library(lmerTest)
library(emmeans) # the workhorse for inference
library(sf)

# graphing packages
library(ggsci) # color palettes
library(ggpubr) # publication quality plots
library(ggforce) # better jitter
library(cowplot) # combine plots
library(knitr) # kable tables
library(kableExtra) # kable_styling tables
library(jpeg)

# ggplot_the_model.R packages not loaded above
library(insight)
library(afex) # aov4

# use here from the here package
here <- here::here
# use clean_names from the janitor package
clean_names <- janitor::clean_names
# use transpose from data.table
transpose <- data.table::transpose

# load functions used by this text written by me
# ggplot_the_model.R needs to be in the folder "R"
# if you didn't download this and add to your R folder in your
# project, then this line will cause an error

source(here("code", "ggplot_the_model.R"))
source(here("code", "ggptm.R"))
source(here("code", "superimposition.R"))

data_folder <- "data"

# constants
dial_to_speed <- 6.42237212 # from my smoke test with screens and forcing through origin

```

# Selected lines have continued to increase mean apparent flying speed for 30+ years

Selection started on or before 1991 and continued until 2023.

```{r}
#| label: import-historic-wind-tunnel-data
#| echo: false
#| warning: false
#| eval: true

subfolder <- "wind tunnel"
filename <- "CURRENT WIND TUNNEL GENERATION MEANS.xlsx"
in_path <- here(data_folder, subfolder, filename)

wt_wide <- read_xlsx(in_path,
                sheet = "Sheet1",
                range = "A1:E443") |>
#  clean_names() |>
  data.table()
wt_wide <- wt_wide[gen != ""]
wt <- melt(wt_wide,
           id.vars = "gen",
           measure.vars = c("AA1-1",
                            "AA1-2",
                            "AA2-1",
                            "AA2-2"),
           variable.name = "id",
           value.name = "mafs")
wt[, line := substr(id,1,3)]
wt <- na.omit(wt)

# compute line means (in some generations there were two runs per line)
line_means <- wt[, .(mafs = mean(mafs)),
                 by = .(gen, line)]
aa_mean <- line_means[, .(mafs = mean(mafs)),
                 by = .(gen)]

```

```{r}
#| label: plot-historic-wt-data
#| echo: false
#| warning: false
#| eval: true

gg <- ggplot(data = wt,
             aes(x = gen,
                 y = mafs,
                 color = line)) +
  geom_point(size = 0.5) +
  
  # each AA line is separate
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 20),
    method.args = list(method = "REML"),
    se = FALSE
  ) +
  
  # Combine aa lines
  # geom_smooth(
  #   data = aa_mean,
  #   aes(x = gen, y = mafs),
  #   method = "gam",
  #   formula = y ~ s(x, k = 10),
  #   method.args = list(method = "REML"),
  #   se = FALSE,
  #   inherit.aes = FALSE,
  #   color = "black",
  #   linewidth = 1.2
  # ) +

  
  scale_color_manual(values = pal_okabe_ito) +
  xlab("Generation") +
  ylab("Mean Apparent Flying Speed (cm/s)") +
  theme_pubr()
gg
```

Concerns:

The AA1 and AA2 lines have remarkably similar fit, which could result from

1. Climate (temperature/humidity) fluctuations in the room during selection runs
2. Climate (temperature/humidity/light) fluctions in the rearing room. Since 2001 the temperature and humidity have been well controlled.
3. Husbandry modifications
4. Gene flow between lines

```{r}
#| label: check-increase-in-aa
#| echo: false
#| warning: false
#| eval: false

late150 <- line_means[gen >= 150]
late150[, gen100 := (gen - 150) / 100]

m_150 <- gls(
  mafs ~ gen100 + line,
  data = late150,
  correlation = corCAR1(form = ~ gen | line),
  method = "REML"
)
summary(m_150)
intervals(m_150)

late250 <- line_means[gen >= 250]
late250[, gen100 := (gen - 250) / 100]

m_250 <- gls(
  mafs ~ gen100 + line,
  data = late250,
  correlation = corCAR1(form = ~ gen | line),
  method = "REML"
)
summary(m_250)
intervals(m_250)
```

# Selected lines have 30-40% higher upwind flight ability than control or wild type lines
## Experiment 1 (Simons-Walker data)

```{r }
#| label: import-walker-burst-data
#| echo: false
#| warning: false
#| eval: true

subfolder <- "burst speed/simons-walker "
filename <- "Wind Tunnel minus days>6.xlsx"
in_path <- here(data_folder, subfolder, filename)

exp1 <- read_xlsx(in_path,
                    range = "A1:H108") |>
  clean_names() |>
  data.table()

exp1[, id := .I]
exp1[, treatment := factor(treatment, levels = c("CN", "AA"))]
exp1[, sex := factor(sex, levels = c("F", "M"))]
exp1[, line := paste0(treatment, line) |>
       factor(levels = c("CN1", "CN2", "AA1", "AA2"))]
exp1[, wind_speed := dial_50 * dial_to_speed]
exp1[, group := ifelse(treatment == "AA", "Selection", "No Selection") |> factor(levels=c("No Selection", "Selection"))]
```

```{r exp1-checks}
#| label: exp1-max-check
#| echo: false
#| warning: false
#| eval: false

exp1[, .(max = max(wind_speed)), by = .(line, sex)]
```

```{r}
#| label: exp1-init-plot
#| echo: false
#| warning: false
#| eval: false

ggplot(data = exp1,
       aes(x = line,
           y = wind_speed)) +
  geom_jitter(width = 0.1)
```


```{r }
#| label: exp1-means-check
#| echo: false
#| warning: false
#| eval: false

exp1_summary <- exp1[, .(wind_speed = mean(wind_speed)), by = .(treatment, sex, line)]
exp1_summary
```

### Linear mixed model

```{r}
#| label: exp1-lmm-random-intercept
#| echo: true
#| warning: false
#| eval: true

exp1_lmm <- lmer(wind_speed ~ treatment * sex + (1 | line), data = exp1)
```

```{r }
#| label: exp1-lmm-stats
#| echo: false
#| warning: false
#| eval: true

exp1_sigma <- summary(exp1_lmm)$sigma

exp1_emm <- emmeans(exp1_lmm, specs = c("treatment", "sex"))

cn_f <- c(1,0,0,0)
aa_f <- c(0,1,0,0)
cn_m <- c(0,0,1,0)
aa_m <- c(0,0,0,1)
exp1_contrast <- list(
  list("F: AA - CN" = aa_f - cn_f),
  list("M: AA - CN" = aa_m - cn_m)
)

exp1_pairs <- contrast(exp1_emm,
                      method = exp1_contrast,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
# standardized effect
exp1_pairs[, effect_std := estimate / exp1_sigma]

# percent effect
exp1_cn_means <- summary(exp1_emm)[c(1,3), "emmean"]
exp1_pairs[, effect_perc := estimate / exp1_cn_means * 100]

# reorder for plot
exp1_pairs[, contrast := factor(contrast,
                                levels = c("M: AA - CN",
                                           "F: AA - CN"))]
```


```{r }
#| label: exp1-lmm-emm
#| echo: false
#| warning: false
#| eval: true
exp1_emm |>
  summary() |>
  data.table() |>
  kable(digits = c(1,1,1,2,1,1,1)) |>
  kable_styling()
```


```{r }
#| label: exp1-lmm-pairs
#| echo: false
#| warning: false
#| eval: true
exp1_pairs |>
  kable(digits = c(1,1,2,1,1,1,1,3,1)) |>
  kable_styling()
```


```{r exp1-conditional-means}
#| label: exp1-conditional-means
#| echo: false
#| warning: false
#| eval: true

# compute conditional means
newdat <- expand.grid(
  line = levels(exp1$line),
  sex = levels(exp1$sex)
) |>
  data.table()
newdat[, treatment := ifelse(substr(newdat$line, 1, 2) == "AA", "AA", "CN")]
newdat[, fit := predict(exp1_lmm, newdata = newdat)]

# check with my own math
coefs <- coef(exp1_lmm)$line
mm_females <- matrix(c(1,0,0,0,
                       1,0,0,0,
                       1,1,0,0,
                       1,1,0,0), nrow=4)
mm_males <- matrix(c(1,0,1,0,
                       1,0,1,0,
                       1,1,1,1,
                       1,1,1,1), nrow=4)
male_means <- as.matrix(coefs) %*% mm_males |> diag()
female_means <- as.matrix(coefs) %*% mm_females |> diag()

# compare with raw means
fixed_means <- exp1[, .(speed = mean(wind_speed)),
                      by = .(sex, line, treatment)]
# fixed_means

```

```{r }
#| label: exp1-conditional-mean-plot
#| echo: false
#| warning: false
#| eval: true

means_plot <- function(plot_sex = "F", title = "Female"){
  offset <- 0.15
  x_breaks <- c(1-offset, 1+offset, 2-offset, 2+offset)
  
  plot_data <- exp1[, .SD, .SDcols = c("treatment", "line", "sex", "wind_speed")]
  treatment_means <- exp1_emm |> 
    summary() |>
    data.table()
  treatment_means[treatment == "CN", x := 1]
  treatment_means[treatment == "AA", x := 2]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "CN1", x := 1 - offset]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "AA1", x := 2 - offset]
  plot_data[line == "AA2", x := 2 + offset]
  newdat[line == "CN1", x := 1 - offset]
  newdat[line == "CN2", x := 1 + offset]
  newdat[line == "AA1", x := 2 - offset]
  newdat[line == "AA2", x := 2 + offset]
  
  gg <- ggplot(data = plot_data[sex == plot_sex],
               aes(x = x,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    geom_jitter(width = 0.05,
                height = 0,
                color = "gray") +
    
    # line means
    geom_point(data = newdat[sex == plot_sex],
               aes(x = x,
                   y = fit),
               size = 2) +
    
    # treatment means and CI
    geom_point(data = treatment_means[sex == plot_sex],
               aes(x = x,
                   y = emmean),
               size = 3) +
    geom_errorbar(data = treatment_means[sex == plot_sex],
                  aes(x = x,
                      y = emmean,
                      ymin = lower.CL,
                      ymax = upper.CL),
                  width = 0) +
    
    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = pal_okabe_ito) +
    scale_x_continuous(breaks = x_breaks,
                       labels = c("CN1", "CN2", "AA1", "AA2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(90, 210),
                    xlim = c(0.7, 2.3)) +
    ggtitle(title)
  return(gg)
}

gg_f <- means_plot(plot_sex = "F", title = "Female")
gg_m <- means_plot(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```

### Supplementary models (in supplement)
#### Classical ANCOVA

```{r}
#| label: exp1-lm-room-temp
#| echo: true
#| warning: false
#| eval: true

lm_ancova <- lm(wind_speed ~ treatment * sex + room_temp, data = exp1)

```

```{r }
#| label: exp1-lm-room-temp-stats
#| echo: false
#| warning: false
#| eval: true

coef(summary(lm_ancova))

ggplot(data = exp1,
       aes(x = room_temp,
           y = wind_speed,
           color = treatment)) +
  geom_point() +
#  geom_ancova(lm_ancova) +
  NULL
```

#### Random intercept with covariate

```{r }
#| label: exp1-lmm-room-temp
#| echo: true
#| warning: false
#| eval: true

exp1_lmm_2 <- lmer(wind_speed ~ treatment * sex + room_temp + (1 | line), data = exp1)
```


```{r }
#| label: exp1-lmm-room-temp-stats
#| echo: false
#| warning: false
#| eval: true

sigma <- summary(exp1_lmm_2)$sigma

coef_table <- coef(summary(exp1_lmm_2))
colnames(coef_table)[5] <- "p-value"
coef_dt <- data.table(
  coefficient = row.names(coef_table),
  coef_table
)
coef_table[2:5,] |>
  kable(digits = c(1,2,1,1,3)) |>
  kable_styling()


exp1_lmm_2_emm <- emmeans(exp1_lmm_2, specs = c("treatment", "sex"))

exp1_lmm_2_pairs <- contrast(exp1_lmm_2_emm,
                      method = exp1_contrast,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
exp1_lmm_2_pairs[, effect_std := estimate / sigma]

exp1_lmm_2_pairs|>
  kable(digits = c(1,1,2,1,1,1,1,3,1)) |>
  kable_styling()
```

#### Classical nested ANOVA

```{r }
#| label: exp1-classical-nested-anova
#| echo: true
#| warning: false
#| eval: true

exp1_aov <- aov(wind_speed ~ treatment * sex + Error(line / sex),
          data = exp1)
```


```{r }
#| label: exp1-classical-nested-anova-stats
#| echo: true
#| warning: false
#| eval: true
 
summary(exp1_aov)

```

## Experiment 2 (Wheeler) data

```{r }
#| label: exp2-clean-his-files
#| echo: false
#| warning: false
#| eval: false
 
# this chunk is to check and clean all of Jess' excel files, which had duplicate entries
subfolder <- "burst speed/jess wheeler/sent from jess/raw"
filename <- "02.14.10.xlsx"
in_path <- here(data_folder, subfolder, filename)

  file_i <- read_xlsx(in_path,
                    range = "a1:c1000") |>
    clean_names() |>
    data.table() |>
    na.omit()
  # angle is flight angle in degrees from 1,0
  # length is flight distance over 1 frame
  colnames(file_i)[1] <- "id"
  file_i[, treatment := substr(id,1,2)]
  file_i[, line := substr(id,1,3)]
  file_i[, sex := substr(id, 4,4)]
  file_i[, dial := substr(id, 5,6) |> as.integer()]
  file_i[, date := substr(id, 8,15)]
 
  check_n <- file_i[, .(N = .N),
     by = .(date, sex, treatment, line, id)]
  check_n[N >=6, id]

  check_id <- check_n[, .(N = .N), 
     by = .(date, sex, treatment, line)]
setorder(check_id, sex, treatment, line, date)

```

```{r }
#| label: exp2-import-cleaned
#| echo: false
#| warning: false
#| eval: true
 
subfolder <- "burst speed/jess wheeler/sent from jess/raw"
subfolder_path <- here(data_folder, subfolder)

files <- list.files(
  path = subfolder_path,
  pattern = "\\.xlsx$",
  full.names = TRUE
)

exp2 <- data.table(NULL)

for(i in 1:length(files)){
  in_path <- files[i]
  
  file_i <- read_xlsx(in_path,
                    range = "a1:c1000") |>
    clean_names() |>
    data.table() |>
    na.omit()
  # angle is flight angle in degrees from 1,0
  # length is flight distance over 1 frame
  colnames(file_i)[1] <- "id"
  file_i[, treatment := substr(id,1,2)]
  file_i[, line := substr(id,1,3) |> 
           toupper() |>
           factor()]
  file_i[, sex := substr(id, 4,4) |> 
           toupper() |>
           factor()]
  file_i[, dial := substr(id, 5,6) |> as.integer()]
  file_i[, date := substr(id, 8,15)]
  
  # performance measures
  file_i[, theta := ifelse(angle < 0,
                         -(angle + 180),
                         180-angle)]
  file_i[, wind_jess := 7.3888 * dial]
  file_i[, wind_jaw := 6.42237212 * dial]
  file_i[, disp.x := length * cos(theta * pi/180)]
  file_i[, disp.y := length * sin(theta * pi/180)]
  file_i[, wind.x := wind_jess / 30]
  file_i[, vel := sqrt((wind.x + disp.x)^2 + disp.y^2) * 30]
  
  exp2 <- rbind(exp2, file_i)
  
}

  # clean treatments and make factor
  exp2[treatment == "aa", treatment := "AA"]
  exp2[treatment == "cn", treatment := "CN"]
  # exp2[treatment == "k4", treatment := "K45"]
  # exp2[treatment == "n6", treatment := "N60"]
  exp2[treatment == "k4", treatment := "Wild"]
  exp2[treatment == "n6", treatment := "Wild"]
  exp2[, treatment := factor(treatment, levels = 
                                 c("CN", "Wild", "AA"))]
  exp2[, line := factor(line, 
                        levels = c("CN1", "CN2",
                                   "K45", "N60",
                                   "AA1", "AA2"))]
  exp2[treatment == "AA", group := "Selection"]
  exp2[treatment == "CN", group := "No Selection"]
  exp2[treatment == "Wild", group := "Wild"]
  exp2[, group := factor(group,
                         levels = c("No Selection",
                                    "Selection",
                                    "Wild"))]

```

```{r }
#| label: exp2-summary
#| echo: false
#| warning: false
#| eval: false
 
# this is just to compare counts with Jess' thesis table
# first compute max within each id
exp2_max_id <- exp2[, .(wind_max = mean(wind_jess),
                        vel_max = mean(vel),
                        N = .N),
     by = .(date, treatment, line, sex, id)]

exp2_sum <- exp2_max_id[, .(wind_max = mean(wind_max),
                            vel_max = mean(vel_max),
                    N = .N), 
     by = .(date, sex, treatment, line)]
setorder(exp2_sum, sex, treatment, line, date)
```

To compare to Exp1 (Simons-Walker) results, I am using wind tunnel speed and not the computed speed of the fly. But I will analyze that in a supplement. Jess has 3 measures per fly; since these are at same tunnel speed, I just need to obtain the summary data as the primary data set.

```{r }
#| label: jess-primary-data
#| echo: false
#| warning: false
#| eval: true

exp2_primary <- exp2[, .(wind_speed = mean(wind_jess)), 
                     by = .(id, treatment, line, sex, group)]
```

### Estimate Treatment Effects

```{r }
#| label: exp2-lmm-random-intercept
#| echo: false
#| warning: false
#| eval: true

exp2_lmm <- lmer(wind_speed ~ treatment * sex + (1 | line), data = exp2_primary)
coef(summary(exp2_lmm))
exp2_sigma <- summary(exp2_lmm)$sigma

exp2_emm <- emmeans(exp2_lmm, specs = c("treatment", "sex"))

cn_f <- c(1,0,0,0,0,0)
wild_f <- c(0,1,0,0,0,0)
aa_f <- c(0,0,1,0,0,0)
cn_m <- c(0,0,0,1,0,0)
wild_m <- c(0,0,0,0,1,0)
aa_m <- c(0,0,0,0,0,1)

# contrasts of interest - this is all contrasts, just not m v f
# aa - cn # primary contrast
# aa - k45 # secondary contrast testing comparison with newer line
# aa - n60 # secondary contrast testing comparison with newer line
# cn - k45 # secondary contrast testing loss in cn
# cn - n60 # secondary contrast testing loss in cn
# n60 - k45 # n60 is july 2009 and k45 is july 2008

exp2_contrasts <- list(
  list("F: AA - CN" = aa_f - cn_f),
  list("F: AA - WT" = aa_f - wild_f),
  list("F: CN - WT" = cn_f - wild_f),
  list("M: AA - CN" = aa_m - cn_m),
  list("M: AA - WT" = aa_m - wild_m),
  list("M: CN - WT" = cn_m - wild_m)
)

exp2_pairs <- contrast(exp2_emm,
                      method = exp2_contrasts,
                      adjust = "none") |>
  summary(infer = TRUE) |>
  data.table()
exp2_pairs[, effect_std := estimate / sigma]
# set up order for effects plot
exp2_pairs[, contrast := factor(contrast,
                                levels = c("M: CN - WT",
                                           "M: AA - WT",
                                           "M: AA - CN",
                                           "F: CN - WT",
                                           "F: AA - WT",
                                           "F: AA - CN"))]

exp2_emm
exp2_pairs

```

```{r }
#| label: exp2-conditional-means
#| echo: false
#| warning: false
#| eval: true

# compute conditional means
newdat <- expand.grid(
  line = levels(exp2_primary$line),
  sex = levels(exp2_primary$sex)
) |>
  data.table()
# newdat[, treatment := line]
# newdat[substr(line,1,2) == "aa", treatment := "aa"]
# newdat[substr(line,1,2) == "cn", treatment := "cn"]
newdat[, treatment := substr(line,1,2)]
newdat[treatment == "K4", treatment := "Wild"]
newdat[treatment == "N6", treatment := "Wild"]
newdat[, fit := predict(exp2_lmm, newdata = newdat)]

```

```{r }
#| label: exp2-conditional-mean-plot
#| echo: false
#| warning: false
#| eval: true

means_plot_dots <- function(plot_sex = "F", title = "Female"){
  offset <- 0.25
  x_breaks <- c(1-offset, 1+offset, 2-offset, 2+offset,
                3-offset, 3+offset)
  
  plot_data <- exp2_primary[, .SD, .SDcols = c("treatment", "line", "sex", "wind_speed")]
  treatment_means <- exp2_emm |> 
    summary() |>
    data.table()
  treatment_means[treatment == "CN", x := 1]
  treatment_means[treatment == "Wild", x := 2]
  treatment_means[treatment == "AA", x := 3]
  plot_data[line == "CN1", x := 1 - offset]
  plot_data[line == "CN2", x := 1 + offset]
  plot_data[line == "K45", x := 2 - offset]
  plot_data[line == "N60", x := 2 + offset]
  plot_data[line == "AA1", x := 3 - offset]
  plot_data[line == "AA2", x := 3 + offset]
  newdat[line == "CN1", x := 1 - offset]
  newdat[line == "CN2", x := 1 + offset]
  newdat[line == "K45", x := 2 - offset]
  newdat[line == "N60", x := 2 + offset]
  newdat[line == "AA1", x := 3 - offset]
  newdat[line == "AA2", x := 3 + offset]

  gg <- ggplot(data = plot_data[sex == plot_sex],
               aes(x = x,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    geom_jitter(width = 0.05,
                height = 0.5,
                size = 0.2,
                color = "gray") +

    # line means
    geom_point(data = newdat[sex == plot_sex],
               aes(x = x,
                   y = fit),
               size = 2) +
    
    # treatment means and CI
    geom_point(data = treatment_means[sex == plot_sex],
               aes(x = x,
                   y = emmean),
               size = 3) +
    geom_errorbar(data = treatment_means[sex == plot_sex],
                  aes(x = x,
                      y = emmean,
                      ymin = lower.CL,
                      ymax = upper.CL),
                  width = 0) +
    
    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = c(pal_okabe_ito[1],
                                  pal_okabe_ito[3],
                                  pal_okabe_ito[2])) +
    scale_x_continuous(breaks = x_breaks,
                       labels = c("CN\n1", "CN\n2",
                                  "Wild\nK45", "Wild\nN60",
                                   "AA\n1", "AA\n2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(0, 250),
                    xlim = c(0.7, 3.3)) +
    ggtitle(title)
  return(gg)
}

gg_f <- means_plot_dots(plot_sex = "F", title = "Female")
gg_m <- means_plot_dots(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```

```{r }
#| label: exp2-conditional-mean-plot-boxplot
#| echo: false
#| warning: false
#| eval: true

means_plot_violin <- function(plot_sex = "F", title = "Female"){

  gg <- ggplot(data = exp2_primary[sex == plot_sex],
               aes(x = line,
                   y = wind_speed,
                   color = treatment)) +
    
    # raw data
    # geom_sina(maxwidth = 0.5,
    #           color = "gray") +
    # geom_jitter(width = 0.05,
    #             height = 0,
    #             color = "gray") +

#    geom_violin() +
    
    geom_boxplot() +


    # clean plot
    ylab("Max Opposing Wind Speed (cm/s)") +
    scale_color_manual(values = 
                         c(pal_okabe_ito[1], 
                           pal_okabe_ito[3], 
                           pal_okabe_ito[2])) +
    scale_x_discrete(labels = c("CN\n1", "CN\n2",
                                "Wild\nK45", "Wild\nN60",
                                "AA\n1", "AA\n2")) +
    theme_pubr() +
    theme(legend.position="none",
          axis.title.x = element_blank()) +
    coord_cartesian(ylim = c(25, 225)) +
    ggtitle(title)
  return(gg)
}

gg_f <- means_plot_violin(plot_sex = "F", title = "Female")
gg_m <- means_plot_violin(plot_sex = "M", title = "Male")
plot_grid(gg_f, gg_m, ncol = 2)
```

## The combined plot suggests that Experiment 2 collected substantial submaximal data

```{r }
#| label: burst-combined-plot
#| echo: false
#| warning: false
#| eval: true

burst1 <- data.table(
  experiment = "Exp1",
  exp1[, .SD, .SDcols = c("treatment", "line", "sex", "group",  "wind_speed")])
burst2 <- data.table(
  experiment = "Exp2",
  exp2[, .SD, .SDcols = c("treatment", "line", "sex", "group", "wind_jess")])
setnames(burst2, old = "wind_jess", new = "wind_speed")
burst <- rbind(burst1, burst2)
burst[, exp_line := paste(experiment, line, sep="\n")]
burst[, exp_line := factor(exp_line, levels = c(
  "Exp1\nCN1", "Exp1\nCN2",
  "Exp1\nAA1", "Exp1\nAA2",
  "Exp2\nCN1", "Exp2\nCN2",
  "Exp2\nAA1", "Exp2\nAA2",
  "Exp2\nK45", "Exp2\nN60"
))]

combined_plot <- function(sex_in = "F", title = "Female"){
  gg <- ggplot(data = burst[sex == sex_in],
               aes(x = exp_line,
                   y = wind_speed,
                   color = group)) +
  geom_jitter(width = 0.2,
              height = 1,
              size = .2) +
  # clean plot
  ylab("Max Opposing Wind Speed (cm/s)") +
  scale_color_manual(values = pal_okabe_ito) +
  theme_pubr() +
  theme(legend.position="none",
        axis.title.x = element_blank()) +
  ggtitle(title)
  return(gg)
}

gg_f <- combined_plot(sex = "F", title = "Female")
gg_f
gg_m <- combined_plot(sex = "M", title = "Male")
gg_m
```

# Head and Thorax
# Wingshape analysis

For wing anatomy, see https://flybase.org/reports/FBim0000853?utm_source=chatgpt.com

and especially
https://pmc.ncbi.nlm.nih.gov/articles/PMC11457947/?utm_source=chatgpt.com

```{r }
#| label: wing-metrics
#| echo: false
#| warning: false
#| eval: true

wing_metrics <- function(x, y, hinge = 1, tip = 17) {

  ## ------------------------------------------------------------
  ## 1. Translate so hinge (LM1) is at the origin
  ## ------------------------------------------------------------

  x0 <- x - x[hinge]
  y0 <- y - y[hinge]


  ## ------------------------------------------------------------
  ## 2. Rotate so hinge -> tip (LM17) is the radial x-axis
  ## ------------------------------------------------------------

  theta <- atan2(y0[tip], x0[tip])

  xr <-  x0 * cos(theta) + y0 * sin(theta)
  yr <- -x0 * sin(theta) + y0 * cos(theta)

  ## Wing length: hinge -> tip
  R <- sqrt(x0[tip]^2 + y0[tip]^2)


  ## ------------------------------------------------------------
  ## 3. Polygon integrals
  ## ------------------------------------------------------------

  # next vertex around perimeter
  x2 <- c(xr[-1], xr[1])
  y2 <- c(yr[-1], yr[1])

  cross <- xr * y2 - x2 * yr

  # signed area
  As <- sum(cross) / 2

  # make formulas independent of clockwise/counterclockwise digitizing
  s <- sign(As)
  A <- abs(As)

  # centroid
  xbar <- sum((xr + x2) * cross) / (6 * As)
  ybar <- sum((yr + y2) * cross) / (6 * As)

  # second moments of AREA about hinge-defined axes
  #
  # Ixx = integral y^2 dA   (chordwise distribution)
  # Iyy = integral x^2 dA   (radial distribution)
  #
  Ixx <- s * sum(
    (yr^2 + yr*y2 + y2^2) * cross
  ) / 12

  Iyy <- s * sum(
    (xr^2 + xr*x2 + x2^2) * cross
  ) / 12

  # product moment
  Ixy <- s * sum(
    (2*xr*yr + xr*y2 + x2*yr + 2*x2*y2) * cross
  ) / 24


  ## ------------------------------------------------------------
  ## 4. Derived wing metrics
  ## ------------------------------------------------------------

  # Ray-style / single-wing aspect ratio convention
  AR <- R^2 / A

  # mean chord
  mean_chord <- A / R

  # radial position of area centroid
  xbar_hat <- xbar / R

  # radius of second moment of area
  r2 <- sqrt(Iyy / A)

  # nondimensional radius of second moment
  r2_hat <- r2 / R
  
  # nondimensional second moment of area
  r2_hat_sq <- Iyy / (A * R^2)


  ## ------------------------------------------------------------
  ## 5. Return results
  ## ------------------------------------------------------------

  data.table(
    area       = A,
    wing_length = R,
    mean_chord = mean_chord,
    aspect_ratio = AR,

    xbar       = xbar,
    ybar       = ybar,
    xbar_hat   = xbar_hat,

    Ixx        = Ixx,
    Iyy        = Iyy,
    Ixy        = Ixy,

    r2         = r2,
    r2_hat     = r2_hat,
    r2_hat_sq  = r2_hat_sq
  )
}

```

```{r }
#| label: radial-area-distribution
#| echo: false
#| warning: false
#| eval: true

# this function is to compare areas along the span vs. Chicoine/Harmon polygons. Harmon found AA flies have smaller proximal polygon but the radial analysis and especially the superimposition simply shows that the main part of the wing starts more distally in AA.

radial_area_distribution <- function(x, y, hinge = 1, tip = 17,
                                     q = seq(0, 1, by = 0.05)) {

  ## Translate hinge to origin
  x0 <- x - x[hinge]
  y0 <- y - y[hinge]

  ## Rotate hinge -> tip onto x-axis
  theta <- atan2(y0[tip], x0[tip])

  xr <-  x0 * cos(theta) + y0 * sin(theta)
  yr <- -x0 * sin(theta) + y0 * cos(theta)

  ## Wing length
  R <- sqrt(x0[tip]^2 + y0[tip]^2)

  ## Close polygon for sf
  xy <- cbind(xr, yr)
  xy <- rbind(xy, xy[1, ])

  wing <- st_polygon(list(xy))

  total_area <- as.numeric(st_area(wing))

  ## Bounds safely beyond wing
  ymin <- min(yr) - R
  ymax <- max(yr) + R
  xmin <- min(xr) - R

  ## For each q, calculate area with x/R <= q
  cumulative_area <- sapply(q, function(qi) {

    xmax <- qi * R

    box <- st_polygon(list(matrix(
      c(xmin, ymin,
        xmax, ymin,
        xmax, ymax,
        xmin, ymax,
        xmin, ymin),
      ncol = 2,
      byrow = TRUE
    )))

    intersection <- suppressWarnings(
      st_intersection(wing, box)
    )

    if (length(intersection) == 0) {
      0
    } else {
      as.numeric(st_area(intersection)) / total_area
    }
  })

  data.table(
    q = q,
    cumulative_area = cumulative_area
  )
}
```

```{r }
#| label: check-daves-regional-area
#| echo: false
#| warning: false
#| eval: true

polygon_area <- function(x, y) {
  
  # next point around polygon, including last -> first
  x2 <- c(x[-1], x[1])
  y2 <- c(y[-1], y[1])
  
  abs(sum(x * y2 - x2 * y) / 2)
}


regional_areas <- function(x, y) {
  
  # Total wing
  total <- polygon_area(x, y)
  
  # Proximal: 1 -> 2 -> 3 -> 4 -> 5 -> 37 -> 1
  prox_ind <- c(1:5, 37)
  proximal <- polygon_area(x[prox_ind], y[prox_ind])
  
  # Middle: 5 -> ... -> 12 -> 27 -> ... -> 37 -> 5
  mid_ind <- c(5:12, 27:37)
  middle <- polygon_area(x[mid_ind], y[mid_ind])
  
  # Distal: 12 -> ... -> 27 -> 12
  dist_ind <- 12:27
  distal <- polygon_area(x[dist_ind], y[dist_ind])
  
  data.table(
    area_total = total,
    area_prox = proximal,
    area_middle = middle,
    area_distal = distal,
    
    prop_prox = proximal / total,
    prop_middle = middle / total,
    prop_distal = distal / total,
    
    partition_error =
      (proximal + middle + distal) - total
  )
}
```


```{r }
#| label: import-wing-data
#| echo: false
#| warning: false
#| eval: true

list_in <- "data/wing shape/wing_files_list.txt"
list_path <- here(list_in)
folder_list <- fread(list_path, sep = "\t",
                     header=FALSE)[[1]]

aero_labels <- c("area", "wing_length","mean_chord", "aspect_ratio", "xbar", "ybar", "xbar_hat", "Ixx", "Iyy", "Ixy", "r2", "r2_hat", "r2_hat_sq")
radii <- c(.1, 0.2, 0.3)
area_labels <- paste0("q", radii)
area_prop_labels <- c("prop_prox")
measure_labels <- c(aero_labels, area_labels, area_prop_labels)

wing_lm <- data.table(NULL) # wing landmarks
wing <- data.table(NULL) # wing metrics
for(i in 1:length(folder_list)){
  folder_i <- folder_list[i]
  rel_folder_path <- paste0("data/wing shape/",folder_i)
  folder_path <- here(rel_folder_path)
  files_short <- list.files( # just the file name
    path = folder_path,
    pattern = "\\.txt$",
    full.names = FALSE
  )
  files_long <- list.files( # full address
    path = folder_path,
    pattern = "\\.txt$",
    full.names = TRUE
  )
  n_files <- length(files_long)
  n_measures <- length(measure_labels)
  wing_measures <- matrix(NA, nrow = n_files, ncol = n_measures)
  colnames(wing_measures) <- measure_labels
  fig_mat <- matrix(NA, nrow = n_files, ncol = 37 * 2)
  colnames(fig_mat) <- c(paste0("x", 1:37),
                         paste0("y", 1:37))
  valid <- rep(FALSE, n_files)
  for(j in 1:n_files){
    lines <- readLines(files_long[j], warn = FALSE)
    lines <- lines[nzchar(trimws(lines))]
    n_landmarks <- length(lines) - 1
    
    if(n_landmarks != 37){
      cat(files_short[j], ":", n_landmarks, "landmarks\n")
      next
    }
    
    valid[j] <- TRUE
    
    fig_lm <- fread(files_long[j], nrows = 37,
                    header = FALSE)
    fig_lm[, V2 := -V2]
    
    # save lm to a matrix for the all fly_id in folder
    fig_mat[j,] <- c(unlist(fig_lm[, V1]),
                     unlist(fig_lm[, V2]))
    # compute wing metrics and save for all fly_id in folder
    aero_j <- wing_metrics(x = fig_lm[, V1],
                           y = fig_lm[, V2])
#    radial area distribution to compare to Harmon's thesis
    area_j <- radial_area_distribution(x = fig_lm[, V1],
                                       y = fig_lm[, V2],
                                       q = radii)
    regional_j <- regional_areas(
      x = fig_lm[, V1],
      y = fig_lm[, V2]
    )
    combined_j <- c(unlist(aero_j[1,]),
                    unlist(area_j[,2]),
                    unlist(regional_j[1, "prop_prox"])
    )
    wing_measures[j, ] <- combined_j
    
  }
  # delete empty rows
  # wing_measures <- na.omit(wing_measures)
  # fig_mat <- na.omit(fig_mat)
  files_short   <- files_short[valid]
  fig_mat       <- fig_mat[valid, , drop = FALSE]
  wing_measures <- wing_measures[valid, , drop = FALSE]
  
  # save landmarks
  wing_lm <- rbind(
    wing_lm,
    data.table(
      file = str_remove(files_short, "\\.txt$"),
      temp_id = str_sub(files_short, 2, 2),
      line_id = str_sub(files_short, 3, 4),
      sex = str_sub(files_short, 5, 5),
      exp_id = str_sub(files_short, 6, 6),
      fly_id = str_extract(files_short, "(?<=^.{6})\\d{1,2}"),
      fig_mat
    )
  )
  # save wing metrics
  wing <- rbind(
    wing,
    data.table(
      file = str_remove(files_short, "\\.txt$"),
      temp_id = str_sub(files_short, 2, 2),
      line_id = str_sub(files_short, 3, 4),
      sex = str_sub(files_short, 5, 5),
      exp_id = str_sub(files_short, 6, 6),
      fly_id = str_extract(files_short, "(?<=^.{6})\\d{1,2}"),
      wing_measures
    )
  )
}

wing[, treatment := ifelse(line_id == "A1" |
                             line_id == "A2",
                           "SEL", "WT") |>
       factor(levels = c("WT", "SEL"))]
wing[, temp_id := factor(temp_id)]
wing[, line_id := factor(line_id,
                         levels = c("K4","L4","N6","A1","A2"))]
wing[, exp_id := factor(exp_id,
                         levels = c("A","B"))]
wing[, sex := factor(sex,
                         levels = c("F","M"))]

wing_lm[, treatment := ifelse(line_id == "A1" |
                                line_id == "A2",
                              "SEL", "WT") |>
          factor(levels = c("WT", "SEL"))]
```

## Superimposed mean wing shapes show that Selected flies have more distal wing

```{r }
#| label: plot-superimposed-wings
#| echo: false
#| warning: false
#| eval: true

# get tps coords
  all_cols <- 1:37
  lm_subset <- c(1,3,5,12,17,21,27,37)
  keep_cols <- all_cols
  p <- length(keep_cols)

  lm_cols <- do.call(paste0,
                     expand.grid(c("x","y"), keep_cols))
  lm_mat <- wing_lm[, .SD, .SDcols = lm_cols] |>
    as.matrix()
  lm_array <- matrix_2_array(lm_mat)
  lm_tpr <- tpr(lm_array, 1, 17)

  lm_dt_tpr <- array_2_matrix(lm_tpr) |>
    data.table()
  lm_cols.tpr <- paste0(lm_cols, ".tpr")
  colnames(lm_dt_tpr) <- lm_cols.tpr
  wing_tpr <- cbind(wing_lm, lm_dt_tpr)
  
  lm_means <- wing_tpr[, lapply(.SD, mean),
                         by = .(treatment),
                         .SDcols = c(lm_cols.tpr)]
  lm_mat <- lm_means[, .SD, .SDcols = lm_cols.tpr] |>
    as.matrix()
  lm_tpr <- matrix_2_array(lm_mat)
  gg <- plot_2shapes(lm_tpr, 
                     color_col = rep(c("SEL", "WT"), each = p))
  gg
  

```

```{r }
#| label: plot-superimposed-wings-by-treatment-combi
#| echo: false
#| warning: false
#| eval: true

# treatment x temp x experiment x sex
# 2 x 3 x 2 x 2
lm_means <- wing_tpr[, lapply(.SD, mean),
                    by = .(treatment, temp_id,
                           exp_id, sex),
                    .SDcols = c(lm_cols.tpr)]

design <- CJ(
#  sex     = c("F", "M"),
  sex     = c("F"),
  temp_id = c("X", "Y", "Z"),
  exp_id  = c("A", "B")
)
plots <- vector("list", nrow(design))
for(i in 1:nrow(design)){
  lm_ss <- lm_means[temp_id == design[i, temp_id] &
                      exp_id == design[i, exp_id] &
                      sex == design[i, sex],
                    .SD, .SDcols = c("treatment", lm_cols.tpr)]
  lm_mat <- lm_ss[, .SD, .SDcols = lm_cols.tpr] |>
    as.matrix()
  lm_tpr <- matrix_2_array(lm_mat)
  plots[[i]] <- plot_2shapes(lm_tpr, 
                      color_col = rep(c("SEL", "WT"), each = p)) +
    ggtitle(paste(design[i, temp_id],
                  design[i, exp_id],
                  design[i, sex])) +
    theme(legend.position = "none")
}

gg <- plot_grid(plotlist = plots,
          nrow = 3)

ggsave(
  "wing_shapes.png",
  gg,
  width = 8,
  height = 12,
  dpi = 300
)

gg
```

```{r }
#| label: lmm-model
#| echo: false
#| warning: false
#| eval: true

# lmm
wing_metric_plot <- function(y_col = "area",
                             sex_i = "F"){
  r_form <- paste(y_col,
                  "~ treatment * temp_id * sex * exp_id +",
                  "(1 | line_id)") |>
    as.formula()
  area_lmm <- lmer(r_form, data = wing)
  area_emm <- emmeans(
    area_lmm,
    ~ treatment | temp_id * sex
  )
  
  area_pairs <- contrast(
    area_emm,
    method = "revpairwise"
  ) |>
    summary(infer = TRUE) |>
    data.table()
  
  # compute conditional means
  line_means <- expand.grid(
    line_id = levels(wing$line_id),
    sex = levels(wing$sex),
    temp_id = levels(wing$temp_id),
    exp_id = levels(wing$exp_id),
    treatment = levels(wing$treatment)
  ) |>
    data.table()
  line_means[, del := 0]
  line_means[line_id %in% c("A1", "A2") & treatment == "WT", del := 1]
  line_means[line_id %in% c("K4", "L4", "N6") & treatment == "SEL", del := 1]
  line_means <- newdat[del == 0,]
  line_means[, fit := predict(area_lmm, newdata = newdat)]
  line_means[, x_temp_treat := as.integer(temp_id) +
               (as.integer(treatment)-1.5)/2.5]
  line_means[treatment == "WT", x_temp_line_treat := x_temp_treat +
               (as.integer(line_id)-2)/10]
  line_means[treatment == "SEL", x_temp_line_treat := x_temp_treat +
               (as.integer(line_id)-4.5)/10]
  
  # raw data to plot
  plot_data <- wing[, .SD, 
                    .SDcols = c("temp_id", "treatment", "exp_id",
                                "line_id","sex", y_col)]
  plot_data[, x_temp_treat := as.integer(temp_id) +
              (as.integer(treatment)-1.5)/2.5 +
              rnorm(n=nrow(plot_data), mean = 0, sd = 0.01)]
  plot_data[treatment == "WT", x_temp_line_treat := x_temp_treat +
              (as.integer(line_id)-2)/10]
  plot_data[treatment == "SEL", x_temp_line_treat := x_temp_treat +
              (as.integer(line_id)-4.5)/10]
  # plot_data[, .(x = mean(x)), by = .(temp_id, treatment)]
  
  # emmeans to plot
  plot_emm <- area_emm |>
    summary() |>
    data.table()
  plot_emm[, x := as.integer(temp_id) +
             (as.integer(treatment)-1.5)/20]
  
  gg <- ggplot(data = plot_data[sex == sex_i],
               aes(x = x_temp_line_treat,
                   y = get(y_col),
                   color = treatment)) +
    # show the data
    geom_point() +
    
    # line means
    geom_point(data = line_means[sex == sex_i],
               aes(x = x_temp_line_treat,
                   y = fit,
                   shape = exp_id),
               size = 2,
               color = "black") +
    
    # treatment : temp means
    geom_point(data = plot_emm[sex == sex_i],
               aes(x = x,
                   y = emmean),
               size = 3,
               color = "black") +
    geom_point(data = plot_emm[sex == sex_i],
               aes(x = x,
                   y = emmean,
                   color = treatment),
               size = 2) +
    geom_errorbar(data = plot_emm[sex == sex_i],
                  aes(x = x,
                      y = emmean,
                      ymin = lower.CL,
                      ymax = upper.CL,
                      color = treatment)) +
    
    scale_color_manual(values = pal_okabe_ito) +
    scale_x_continuous(breaks = 1:3,
                       labels = c("23°", "26°", "29°")) +
    coord_cartesian(xlim = c(0.5, 3.5)) +
    theme_pubr() +
    theme(#legend.position="none",
      axis.title.x = element_blank()) +
    NULL
  return(gg)
  
}

gg_area_f <- wing_metric_plot(y_col = "area", sex_i = "F") +
  ylab(expression(paste("Wing Area (", mm^2, ")")))
gg_area_m <- wing_metric_plot(y_col = "area", sex_i = "M") +
  ylab(expression(paste("Wing Area (", mm^2, ")")))
gg_area_f
gg_area_m
gg_AR_f <- wing_metric_plot(y_col = "aspect_ratio", sex_i = "F") +
  ylab("Aspect Ratio")
gg_AR_m <- wing_metric_plot(y_col = "aspect_ratio", sex_i = "M") +
  ylab("Aspect Ratio")
gg_AR_f
gg_AR_m
gg_WL_f <- wing_metric_plot(y_col = "wing_length", sex_i = "F") +
  ylab("Wing Length (mm)")
gg_WL_m <- wing_metric_plot(y_col = "wing_length", sex_i = "M") +
  ylab("Wing Length (mm)")
gg_WL_f
gg_WL_m
gg
```

```{r}
#| label: check-outliers
#| echo: false
#| warning: false
#| eval: false
wing[sex == "F" & wing_length < 2, file]
# "WXN6FB9pp" is a crazy short wing
```

```{r}
percent_change <- function(y_label){
  r_form <- paste0("log(", y_label,
  ") ~ treatment * temp_id * sex * exp_id + (1 | line_id)")
  m1 <- lmer(r_form, data = wing)
  m1_emm <- emmeans(m1,
                    ~ treatment,
                    weights = "equal"
  )
#  m1_pairs <- contrast(m1_emm, method = "revpairwise")
  m1_pairs <- contrast(m1_emm, method = list("SEL - WT" = c(-1, 1)))
  pairs_dt <- m1_pairs |>
    summary() |>
    data.table()
  estimate <- pairs_dt[1, "estimate"]
  change <- 100 * (exp(estimate) - 1)
  return(unlist(c(estimate, change)))
}

change_matrix <- matrix(NA, nrow = 3, ncol = 2)
colnames(change_matrix) <- c("estimate", "percent")
change_matrix[1, ] <- percent_change("area")
change_matrix[2, ] <- percent_change("aspect_ratio")
change_matrix[3, ] <- percent_change("r2_hat_sq")
change_dt <- data.table(
  measure = c("area", "aspect_ratio", "r2_hat_sq"),
  change_matrix
)




```

```{r}
percent_change_2 <- function(y_label){
  r_form <- paste0("log(", y_label,
  ") ~ treatment * temp_id * exp_id * sex + (1 | line_id)")
  m1 <- lmer(r_form, data = wing)
  m1_emm <- emmeans(m1,
                    ~ treatment | temp_id * exp_id * sex
  )
  m1_pairs <- contrast(m1_emm, method = list("Sel - CN" = c(-1, 1)))
  m1_pairs_df <- as.data.frame(m1_pairs)
  m1_pairs_df$percent <- 100 * (exp(m1_pairs_df$estimate) - 1)
  return(m1_pairs_df)
}

keep_cols <- c("temp_id","sex","exp_id","percent")

res <- percent_change_2("area")
change_dt <- res[, keep_cols]
colnames(change_dt)[4] <- "area"

res <- percent_change_2("wing_length")
change_dt$wing_length <- as.data.frame(res)$percent

res <- percent_change_2("mean_chord")
change_dt$mean_chord <- as.data.frame(res)$percent

res <- percent_change_2("aspect_ratio")
change_dt$aspect_ratio <- as.data.frame(res)$percent

res <- percent_change_2("xbar_hat")
change_dt$xbar_hat <- as.data.frame(res)$percent

res <- percent_change_2("r2_hat_sq")
change_dt$r2_hat_sq <- as.data.frame(res)$percent

change_dt
# clipr::write_clip(change_dt)
```
